```{r Setup}
library(tidyverse) #v2.0.0
library(stringr) #v1.6.0
library(ggrastr) #v1.0.2
library(dplyr) #v1.2.1
library(Seurat) #v5.4.0
library(scales) #1.4.0
library(cowplot) #v1.2.0
library(fmsb) #v0.7.6
library(fgsea) #v1.36.0
library(ZemmourLib) #v0.1.1
library(pheatmap) #v1.0.13
library(RColorBrewer) #v1.1-3
library(ggrepel) #v0.9.8
library(msigdbr) #v25.1.1
library(viridis) #v0.6.5
library(viridisLite) #v0.4.3
library(ggpubr) #v0.5.9
library(scattermore) #v1.2
library(patchwork) #v1.3.2
setwd("/Users/nityamehrotra/Library/CloudStorage/Box-Box/ZemmourLab_Nitya/CD4")

cd4_object <- readRDS("cd4_diet.rds")

color_palette <- read.csv("immgent_color_palettes_level2.csv",header = FALSE)

palette_vector <- setNames(as.character(color_palette$V2), as.character(color_palette$V1))

organ_palette <- read.csv("immgent_color_palettes_organ_simplified.csv",header = FALSE)

palette_vector_organ <- setNames(
  as.character(organ_palette$V2),
  as.character(organ_palette$V1)
)

condition_palette <- read.csv("immgent_colors_level1.csv",header = FALSE)

palette_level1 <- setNames(as.character(condition_palette$V2),as.character(condition_palette$V1))

cd4_cite <- subset(cd4_object,subset = cite_seq == "TRUE")

smarta_ids <- read.csv("smarta_citeseq_igt363840_2026.csv",header = FALSE)$V1

tcr <- read.csv("all_igts_summary.csv.gz")

xy_genes <- readLines("XY_vs_RSTUHI_up.txt")
rs_genes <- readLines("RS_vs_TUXYHI_up.txt")
tu_genes <- readLines("TU_vs_RSXYHI_up.txt")
hi_genes <- readLines("HI_vs_RSTUXY_up.txt")
polyfunctional_meta <- read.csv("polyfunctional_metadata.csv")
resting_analysis <- readRDS("ttlist_in_Resting.Rds")
q_dge <- read.csv("Q_program_volcano_DGE_all.csv")
obj <- readRDS("b6spleenb6ln.Rds")
```
# Figure 1

## immgenT resolves CD4+ T cell heterogeneity into 20 discrete clusters and a reference embedding

### Figure 1a

```{r 1a - overall MDE - David}

```
### Figure 1b
```{r 1b - overall CD4 MDE, fig.width=8, fig.height=8}
DimPlot(cd4_object, reduction = "mde_incremental", group.by = "annotation_level2", cols = palette_vector, label = TRUE, raster = TRUE, raster.dpi = c(1024, 1024)) + ggtitle("CD4 T Cells in All Populations") + NoAxes() + NoLegend() + theme(legend.position = "none") + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5))
```
### Figure 1c
```{r 1c - cd62l x cd44 feature plots for resting/activated, fig.width=8, fig.height=8}
cd4_cite <- NormalizeData(cd4_cite, assay = "ADT", normalization.method = "LogNormalize")

resting_clusters <-c("CD4.A", "CD4.B", "CD4.C", "CD4.D", "CD4.E", "CD4.F")
cd4_resting <- subset(cd4_cite, subset = annotation_level2 %in% resting_clusters)
cd4_resting$not_cl6 <- cd4_resting$annotation_level2 != "CD4_cl6"
p <- MyFeatureScatter(
  so = cd4_resting,
  group.by = "not_cl6",
  feature1 = "CD62L",
  feature2 = "CD44",
  raster = TRUE
)
print(p)

activated_clusters <- c("CD4.G", "CD4.H", "CD4.I", "CD4.Q", "CD4.R", "CD4.S", "CD4.T", "CD4.U", "CD4.X", "CD4.Y")
cd4_activated <- subset(cd4_cite, subset = annotation_level2 %in% activated_clusters)
cd4_activated$not_cl6 <- cd4_activated$annotation_level2 != "CD4_cl6"
p <- MyFeatureScatter(
  so = cd4_activated,
  group.by = "not_cl6",
  feature1 = "CD62L",
  feature2 = "CD44",
  raster = TRUE
)
print(p)
```
### Figure 1d
```{r 1d - dotplot of selected markers, fig.width= 4.5, fig.height= 2.5}
gene_markers <- c("Sell", "Cd44", "Bcl6", "Cxcr5", "Tox", "Tox2", "Tigit", "Lag3", "Ctla4", "Havcr2", "Il21", "Pdcd1", "Izumo1r", "Tnfrs4", "Tbx21", "Ifng", "Gzma", "Gzmb", "Gata3", "Il4", "Il5", "Il13", "Rorc", "Il17a", "Il22", "Il10")

DefaultAssay(cd4_object) <- "RNA"

genes_use <- intersect(gene_markers, rownames(cd4_object[["RNA"]]))
cd4_object <- NormalizeData(cd4_object, assay = "RNA", normalization.method = "LogNormalize", layer = "data")

dp <- DotPlot(
  cd4_object,
  features = genes_use,
  group.by = "annotation_level2",
  assay = "RNA"
)

dp_df <- dp$data

remove_ids <- c("CD4.P", "CD4.wM", "CD4.wN", "CD4.wZ")
dp_df <- dp_df %>%
  filter(!id %in% remove_ids)

dp_df$id <- droplevels(dp_df$id)

dp_df$features.plot <- factor(dp_df$features.plot, levels = rev(genes_use))


base_size <- 10
axis_size <- 9

p_rna <- ggplot(dp_df, aes(x = id, y = features.plot)) +
  geom_point(
    aes(size = pct.exp, fill = avg.exp.scaled),
    shape = 21,
    color = "black",
    stroke = 0.12
  ) +
  scale_fill_gradient2(
    low = "darkblue",
    mid = "white",
    high = "red",
    midpoint = 0,
    oob = squish,
    name = "Avg expr (scaled)"
  ) +
  scale_size_continuous(
    name = "% expressing",
    range = c(0.5, 4),
    breaks = c(0, 25, 50, 75, 100),
    limits = c(0, 100)
  ) +
  scale_y_discrete(expand = expansion(mult = c(0.03, 0.03))) +
  scale_x_discrete(expand = expansion(mult = c(0.04, 0.04))) +
  theme_bw(base_size = base_size) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(
      size = axis_size,
      angle = 90,
      vjust = 0.5,
      hjust = 1,
      margin = margin(t = 1)
    ),
    axis.text.y = element_text(
      size = axis_size,
      lineheight = 0.85,
      margin = margin(r = 1)
    ),
    panel.grid = element_blank(),
    panel.border = element_rect(linewidth = 0.12),
    plot.margin = margin(2, 2, 2, 2),
    legend.title = element_text(size = base_size),
    legend.text  = element_text(size = base_size - 1)
  ) +
  guides(
    size = guide_legend(title = "% expressing"),
    fill = guide_colorbar(title = "Avg expr (scaled)")
  )

print(p_rna)
```
### Figure 1e
```{r 1e - condition dotplot and organ dotplot}
cd4_object@meta.data <- cd4_object@meta.data %>%
  mutate(
    condition_detailed_simplified = trimws(as.character(condition_detailed_simplified)),
    condition_detailed            = trimws(as.character(condition_detailed)),
    condition_broad               = trimws(as.character(condition_broad))
  )

cd4_object@meta.data <- cd4_object@meta.data %>%
  mutate(
    condition_detailed_simplified = ifelse(
      condition_detailed %in% c("12mo", "18mo", "27mo"),
      "aging",
      condition_detailed_simplified
    )
  )

cd4_object@meta.data <- cd4_object@meta.data %>%
  mutate(
    condition_detailed_simplified = case_when(
      condition_detailed_simplified %in% c("SFB", "SFB_pregnancy") ~ "SFB",
      TRUE ~ condition_detailed_simplified
    )
  )


cd4_object@meta.data <- cd4_object@meta.data %>%
  mutate(
    condition_detailed_simplified = case_when(
      condition_detailed_simplified %in% c("PDAC", "PDAC_aPD1aCTLA4", "PDAC_FishHFD", "PDAC_SoyHFD") ~ "PDAC",
      condition_detailed_simplified %in% c("GBMclass", "GBMclass_aPD1aCTLA4",
                                           "GBMmes", "GBMmes_aPD1aCTLA4",
                                           "GBMpron", "GBMpron_aPD1aCTLA4") ~ "GBM",
      condition_detailed_simplified %in% c("B16_ACT", "B16_ACTaPD1a41BB") ~ "B16",
      condition_detailed_simplified %in% c("HeartTransplant_rejection",
                                           "HeartTransplant_tolerance") ~ "HeartTransplant",
      TRUE ~ condition_detailed_simplified
    )
  )

cd4_object <- subset(
  cd4_object,
  subset = !condition_detailed_simplified %in% c(
    "MHCIIhet_Ea16", "MHCIIhet",
    "MHCIIKO_Ea16", "MHCIIKO",
    "KbDbKO", "KbDbKO_MCMV",
    "KbDbQa1KO", "KbDbQa1KO_MCMV", "other"
  )
)


cd4_object <- subset(
  cd4_object,
  subset = !(condition_detailed_simplified == "B16" &
             condition_broad == "healthy")
)

sample_counts <- table(cd4_object$sample_name)
keep_samples  <- names(sample_counts[sample_counts > 10])

cd4_filt <- subset(cd4_object, subset = sample_name %in% keep_samples)

cd4_filt@meta.data <- cd4_filt@meta.data %>%
  mutate(
    condition_detailed_simplified = trimws(as.character(condition_detailed_simplified))
  )

cond_col    <- "condition_detailed_simplified"
cluster_col <- "annotation_level2"
broad_col   <- "condition_broad"

base_size <- 6.5
axis_size <- 6

cd4_metadata <- cd4_filt@meta.data

broad_order <- c(
  "healthy", "bacteria", "fungus", "virus", "parasite",
  "multiinfection", "autoimmune", "allergy", "cancer",
  "immunization", "allotransplant"
)

cd4_metadata <- cd4_metadata %>%
  mutate(
    condition_broad = factor(condition_broad, levels = broad_order)
  )

cond_to_broad <- cd4_metadata %>%
  count(
    condition = .data[[cond_col]],
    broad     = .data[[broad_col]],
    name = "cells"
  ) %>%
  group_by(condition) %>%
  slice_max(cells, n = 1, with_ties = FALSE) %>%
  ungroup()

dotplot <- cd4_metadata %>%
  count(
    condition = .data[[cond_col]],
    cluster   = .data[[cluster_col]],
    name = "n"
  ) %>%
  left_join(cond_to_broad, by = "condition") %>%
  filter(!is.na(condition), !is.na(cluster), !is.na(broad)) %>%
  group_by(condition) %>%
  mutate(total = sum(n), prop = n / total) %>%
  ungroup() %>%
  mutate(
    n_log10 = log10(n + 1),
    broad   = factor(broad, levels = broad_order)
  )

cond_levels <- dotplot %>%
  distinct(condition, broad, total) %>%
  arrange(broad, desc(total), condition) %>%
  pull(condition)

dotplot <- dotplot %>%
  mutate(condition = factor(condition, levels = rev(cond_levels)))
side <- dotplot %>% distinct(condition, broad)
broad_levels <- levels(droplevels(dotplot$broad))

broad_cols <- c(
  "lightblue4", "lightblue1", "purple4", "green4", "yellow3",
  "lightyellow3", "pink3", "orange", "pink4", "orange4", "darkred"
)
broad_cols <- broad_cols[seq_len(length(broad_levels))]
names(broad_cols) <- broad_levels


p_side <- ggplot(side, aes(x = 1, y = condition, fill = broad)) +
  geom_tile() +
  scale_fill_manual(values = broad_cols, drop = FALSE) +
  scale_y_discrete(expand = expansion(mult = c(0.01, 0.01))) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = base_size - 1),
    legend.text  = element_text(size = base_size - 2),
    legend.key.height = unit(0.18, "cm"),
    legend.key.width  = unit(0.38, "cm"),
    legend.key = element_rect(color = "black", linewidth = 0.3),
    plot.margin = margin(0, 0, 0, 0)
  ) +
  guides(fill = guide_legend(title = "Condition broad", nrow = 2, byrow = TRUE))

legend_broad <- get_legend(p_side)
p_side_noleg <- p_side +
  theme(legend.position = "none") +
  theme(plot.margin = margin(2, 0, 2, 0))

p_main <- ggplot(dotplot, aes(x = cluster, y = condition)) +
  geom_point(aes(size = n_log10, fill = prop),
             shape = 21, color = "black", stroke = 0.1) +
  scale_fill_gradient(low = "white", high = "darkgreen",
                      limits = c(0, 0.25), oob = squish) +
  scale_size_continuous(name = "Cells per dot (log10)", range = c(1.5, 4)) +
  theme_bw(base_size = base_size) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(size = axis_size - 1, angle = 90, vjust = 0.5, hjust = 1),
    axis.text.y = element_text(size = axis_size - 1, lineheight = 0.75),
    panel.grid = element_blank(),
    panel.border = element_rect(linewidth = 0.1),
    plot.margin = margin(2, 2, 2, 0)
  ) +
  guides(
    size = guide_legend(title = "Cells per dot (log10)"),
    fill = guide_colorbar(title = "Mean proportion")
  )

bottom_row <- plot_grid(
  p_side_noleg, p_main,
  ncol = 2, rel_widths = c(0.02, 0.98),
  align = "h", axis = "tb"
)

plot_grid(legend_broad, bottom_row, ncol = 1, rel_heights = c(0.05, 0.95))

#conditions
cond_order <- rev(levels(dotplot$condition))
organ_order <- c("spleen", "LN", "blood", "bone marrow","lung", "colon","small intestine", "skin", "kidney", "liver", "pancreas", "submandibular gland", "synovial fluid", "placenta", "uterus", "prostate", "mammary gland", "peritoneal cavity","CNS")

org_df <- cd4_filt@meta.data %>%
  mutate(
    condition = trimws(as.character(condition_detailed_simplified)),
    organ     = trimws(as.character(organ_simplified))
  ) %>%
  filter(!is.na(condition), condition != "",
         !is.na(organ), organ != "") %>%
  distinct(condition, organ) %>%
  mutate(
    condition = factor(condition, levels = rev(cond_order)),
    organ     = factor(organ, levels = organ_order)
  ) %>%
  filter(!is.na(condition), !is.na(organ))

p_org_presence <- ggplot(org_df, aes(x = organ, y = condition)) +
  geom_point(
    aes(fill = organ),
    shape = 21,
    size = 2.0,
    color = "black",
    stroke = 0.1
  ) +
  scale_x_discrete(drop = FALSE) + 
  scale_fill_manual(values = palette_vector_organ, drop = FALSE) +
  scale_y_discrete(na.translate = FALSE) +
  theme_bw(base_size = base_size) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(size = axis_size - 1, angle = 90, vjust = 0.5, hjust = 1),
    axis.text.y = element_text(size = axis_size - 1, lineheight = 0.75),
    panel.grid = element_blank(),
    panel.border = element_rect(linewidth = 0.1),
    plot.margin = margin(2, 2, 2, 2)
  )
print(p_org_presence)
```
# Figure 2

## Mapping antigen-specific and clonally expanded cells across immune challenges

### Figure 2a
```{r 2a - ag-specific MDEs for conditions, fig.width=8, fig.height=8}
mde <- as.data.frame(
  Embeddings(
    cd4_object,
    reduction = "mde_incremental"
  )
)

colnames(mde)[1:2] <- c(
  "MDE_1",
  "MDE_2"
)

mde$Cell_ID <- rownames(mde)


ag_plot_info <- list(

  LCMV_SMARTA = list(
    type = "cell_ids",
    ids = smarta_ids,
    conditions = "LCMVarm",
    title = "LCMVarm SMARTA"
  ),

  Listeria_2WTetp = list(
    type = "Ag_spe",
    ag_spe = "2WTetp",
    conditions = "Lmonocytogenes",
    title = "Listeria : 2WTetp"
  ),

  Chlamydia_NR1 = list(
    type = "Ag_spe",
    ag_spe = "NR1",
    conditions = c(
      "CtrachomatisD",
      "CtrachomatisL2"
    ),
    title = "Chlamydia : NR1"
  ),

  HDM_Derp1Tetp = list(
    type = "Ag_spe",
    ag_spe = "Derp1Tetp",
    conditions = "HDM",
    title = "HDM : Derp1Tetp"
  ),

  Toxo_AS15Tetp = list(
    type = "Ag_spe",
    ag_spe = "AS15Tetp",
    conditions = "Toxo",
    title = "Toxo : AS15Tetp"
  )
)


ag_plots <- list()
ag_counts <- list()


for (plot_name in names(ag_plot_info)) {

  info <- ag_plot_info[[plot_name]]

  conditions_use <- info$conditions


  cells_condition <- colnames(cd4_object)[
    !is.na(cd4_object$condition_detailed_simplified) &
      cd4_object$condition_detailed_simplified %in%
        conditions_use
  ]


  if (info$type == "cell_ids") {

    cells_ag <- intersect(
      cells_condition,
      info$ids
    )

  } else if (info$type == "Ag_spe") {

    cells_ag <- colnames(cd4_object)[
      !is.na(cd4_object$Ag_spe) &
        cd4_object$Ag_spe == info$ag_spe &
        !is.na(cd4_object$condition_detailed_simplified) &
        cd4_object$condition_detailed_simplified %in%
          conditions_use
    ]

  }


  cells_condition_non_ag <- setdiff(
    cells_condition,
    cells_ag
  )


  ag_counts[[plot_name]] <- data.frame(
    plot = plot_name,
    n_condition = length(cells_condition),
    n_ag_specific = length(cells_ag),
    n_non_ag_specific = length(cells_condition_non_ag)
  )


  print(
    ag_counts[[plot_name]]
  )


  ag_df <- mde[
    mde$Cell_ID %in% cells_ag,
  ]


  p <- ggplot() +

    geom_scattermore(
      data = mde,
      aes(
        x = MDE_1,
        y = MDE_2
      ),
      color = "lightgray",
      pointsize = 1,
      alpha = 0.5,
      pixels = c(1024, 1024)
    ) +

    geom_scattermore(
      data = mde[
        mde$Cell_ID %in% cells_condition_non_ag,
      ],
      aes(
        x = MDE_1,
        y = MDE_2
      ),
      color = "black",
      pointsize = 2,
      alpha = 1,
      pixels = c(1024, 1024)
    ) +

    geom_point(
      data = ag_df,
      aes(
        x = MDE_1,
        y = MDE_2
      ),
      color = "red",
      size = 1.2,
      alpha = 1
    ) +

    theme_void() +

    ggtitle(
      paste0(
        info$title,
        " (n=",
        length(cells_ag),
        ")"
      )
    ) +

    theme(
      plot.title = element_text(
        size = 10,
        hjust = 0.5
      ),
      legend.position = "none"
    )


  ag_plots[[plot_name]] <- p


  print(p)


}



ag_gallery <- wrap_plots(ag_plots,ncol = 5)

print(ag_gallery)

```
### Figure 2b
```{r 2b plots highlighting cells w/ clonally expanded tcrs (10+) across conditions, colored by clonotype, fig.width=8, fig.height=8}
expanded_threshold <- 9

cd4_tcrs <- tcr %>%
  filter(
    level1 == "CD4",
    !is.na(alpha.vgene),
    !is.na(alpha.jgene),
    !is.na(alpha.junction.nt),
    !is.na(beta.vgene),
    !is.na(beta.jgene),
    !is.na(beta.junction.nt)
  ) %>%
  unite(
    clonotype,
    alpha.vgene,
    alpha.jgene,
    alpha.junction.nt,
    beta.vgene,
    beta.jgene,
    beta.junction.nt,
    sep = ".",
    remove = FALSE
  ) %>%
  mutate(
    annotation_level2 = cd4_object$annotation_level2[
      match(IGT.cellID, colnames(cd4_object))
    ]
  ) %>%
  filter(
    !is.na(annotation_level2),
    !annotation_level2 %in% c("MAIT", "NKT")
  )

tcr_meta <- cd4_tcrs %>%
  select(
    IGT.cellID,
    clonotype
  ) %>%
  distinct(
    IGT.cellID,
    .keep_all = TRUE
  )

cd4_object$clonotype <- tcr_meta$clonotype[
  match(
    colnames(cd4_object),
    tcr_meta$IGT.cellID
  )
]

cells_keep <- colnames(cd4_object)[
  is.na(cd4_object$transgenic_byTCR) |
    !cd4_object$transgenic_byTCR %in% c(
      "2D2",
      "SMARTA",
      "NR1"
    )
]

cd4_plot <- subset(
  cd4_object,
  cells = cells_keep
)

mde <- as.data.frame(
  Embeddings(
    cd4_plot,
    reduction = "mde_incremental"
  )
)

colnames(mde)[1:2] <- c(
  "MDE_1",
  "MDE_2"
)

mde$Cell_ID <- rownames(mde)

meta_df <- cd4_plot@meta.data %>%
  as.data.frame() %>%
  mutate(
    Cell_ID = rownames(.)
  ) %>%
  select(
    Cell_ID,
    condition_detailed_simplified,
    clonotype
  )

mde <- mde %>%
  left_join(
    meta_df,
    by = "Cell_ID"
  ) %>%
  rename(
    condition = condition_detailed_simplified
  )

cat(
  "Cells in MDE:",
  nrow(mde),
  "\n"
)

cat(
  "Cells with clonotypes:",
  sum(!is.na(mde$clonotype)),
  "\n"
)

cat(
  "Cells with conditions:",
  sum(!is.na(mde$condition)),
  "\n"
)

expanded_clonotypes <- mde %>%
  filter(
    !is.na(condition),
    !is.na(clonotype)
  ) %>%
  count(
    condition,
    clonotype,
    name = "n_cells"
  ) %>%
  filter(
    n_cells > expanded_threshold
  )

cat(
  "Expanded clonotypes:",
  nrow(expanded_clonotypes),
  "\n"
)

conditions_to_plot <- sort(
  unique(expanded_clonotypes$condition)
)

cat(
  "Conditions to plot:",
  length(conditions_to_plot),
  "\n"
)

print(conditions_to_plot)

clonotype_gallery <- list()

for (cond in conditions_to_plot) {

  message(
    "Plotting: ",
    cond
  )

  expanded_clones <- expanded_clonotypes %>%
    filter(
      condition == cond
    ) %>%
    arrange(
      desc(n_cells)
    ) %>%
    pull(
      clonotype
    )

  plot_df <- mde %>%
    mutate(
      plot_group = case_when(
        condition == cond &
          clonotype %in% expanded_clones ~ "expanded",

        condition == cond ~ "condition",

        TRUE ~ "background"
      )
    )

  expanded_df <- plot_df %>%
    filter(
      plot_group == "expanded"
    ) %>%
    mutate(
      clonotype = factor(
        clonotype,
        levels = expanded_clones
      )
    )

  clone_colors <- colorRampPalette(
    palette_vector
  )(
    length(expanded_clones)
  )

  names(clone_colors) <- expanded_clones

  p <- ggplot() +

    geom_scattermore(
      data = plot_df %>%
        filter(
          plot_group == "background"
        ),
      aes(
        MDE_1,
        MDE_2
      ),
      color = "lightgray",
      pointsize = 1,
      alpha = 0.35,
      pixels = c(
        1024,
        1024
      )
    ) +

    geom_scattermore(
      data = plot_df %>%
        filter(
          plot_group == "condition"
        ),
      aes(
        MDE_1,
        MDE_2
      ),
      color = "black",
      pointsize = 2,
      alpha = 1,
      pixels = c(
        1024,
        1024
      )
    ) +

    geom_point(
      data = expanded_df,
      aes(
        MDE_1,
        MDE_2,
        color = clonotype
      ),
      size = 1.2,
      alpha = 1
    ) +

    scale_color_manual(
      values = clone_colors
    ) +

    theme_void() +

    ggtitle(
      paste0(
        cond,
        "\n",
        length(expanded_clones),
        " expanded clonotypes"
      )
    ) +

    theme(
      plot.title = element_text(
        size = 9,
        hjust = 0.5
      ),
      legend.position = "none"
    )

  clonotype_gallery[[cond]] <- p

  print(p)
}

if (length(clonotype_gallery) > 0) {

  clonotype_gallery_plot <- wrap_plots(
    clonotype_gallery,
    ncol = 4
  )

  print(
    clonotype_gallery_plot
  )

} else {

  message(
    "No plots generated: no clonotypes passed the expansion threshold."
  )
}
```



# Figure 3

## Activated CD4+ T cells map to large mixed T helper phenotype clusters and smaller polarized clusters
### Fig 3a
```{r 3a MDE highlighting mixed-phenotype clusters (G/Q) and polarized clusters (R/S/T/U/X/Y/H/I), fig.width = 8, fig.height = 8}
gq_clusters <- c("CD4.G", "CD4.Q")

p_gq <- DimPlot(
  cd4_object,
  reduction = "mde_incremental",
  group.by = "annotation_level2",
  cells.highlight = WhichCells(
    cd4_object,
    expression = annotation_level2 %in% gq_clusters
  ),
  cols = "lightgrey",
  cols.highlight = palette_vector[gq_clusters],
  sizes.highlight = 0.6,
  pt.size = 0.2,
  raster = TRUE
) +
  ggtitle("CD4.G / CD4.Q") +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  )

p_gq

# R/S/T/U/X/Y/H/I
polarized_clusters <- c(
  "CD4.R", "CD4.S",
  "CD4.T", "CD4.U",
  "CD4.X", "CD4.Y",
  "CD4.H", "CD4.I"
)

p_polarized <- DimPlot(
  cd4_object,
  reduction = "mde_incremental",
  group.by = "annotation_level2",
  cells.highlight = WhichCells(
    cd4_object,
    expression = annotation_level2 %in% polarized_clusters
  ),
  cols = "lightgrey",
  cols.highlight = palette_vector[polarized_clusters],
  sizes.highlight = 0.6,
  pt.size = 0.2,
  raster = TRUE
) +
  ggtitle("Polarized clusters") +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  )

p_polarized
```

### Fig 3bc
```{r 3b/c feature plots of polarized state TF/cytokine genes, fig.width=8, fig.height=8}
#3b
FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Ifng"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none")

FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Il5"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none")

FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Il17a"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none") 

FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Il21"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none") 

#3c
FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Tbx21"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none")
FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Gata3"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none") 
FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Rorc"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none")

FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Bcl6"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none")
```
### Fig 3d
```{r 3d - mdes highlighting cluster distribution in selected conditions, fig.width = 8, fig.height = 8}
#tgondii
cd4_object$is_toxod9 <- cd4_object$condition_detailed %in% c("Toxo_D9")
cells_toxo <- colnames(cd4_object)[cd4_object$is_toxod9]
plots_toxo <- MyDimPlotHighlight(
  seurat_object         = cd4_object,
  umap_to_plot          = "mde_incremental",
  cells_to_highlight    = cells_toxo,
  highlight_column_name = "annotation_level2",
  title                 = "CD4 Polarized cells highlighted",
  labelclusters = FALSE,
  mycols = palette_vector,
  highlight_size = 0.5,
  highlight_alpha = 1.0,
  background_alpha = 1.0,
  pixels = c(512,512)
)
p <- print(plots_toxo$plot1 + NoAxes() + NoGrid()) + NoLegend() + theme(legend.position = "none") + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5))

print(p)

#tmuris
cd4_object$is_tmuris <- cd4_object$condition_detailed_simplified %in% c("Tmuris")
cells_tmuris <- colnames(cd4_object)[cd4_object$is_tmuris]
plots_tmuris <- MyDimPlotHighlight(
  seurat_object         = cd4_object,
  umap_to_plot          = "mde_incremental",
  cells_to_highlight    = cells_tmuris,
  highlight_column_name = "annotation_level2",
  title                 = "CD4 polarized cells highlighted",
  labelclusters = FALSE,
  mycols = palette_vector,
  highlight_size = 0.5,
  highlight_alpha = 1.0,
  background_alpha = 1.0,
  pixels = c(512,512)
)
p <- print(plots_tmuris$plot1 + NoAxes() + NoGrid()) + NoLegend() + theme(legend.position = "none") + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5))
p
print(p)
#sfb ileum
cd4_object$is_sfb <- cd4_object$condition_detailed_simplified %in% "SFB" & cd4_object$organ == "ileum lamina propria"

cells_is_sfb <- colnames(cd4_object)[cd4_object$is_sfb]
plots_is_sfb <- MyDimPlotHighlight(
  seurat_object         = cd4_object,
  umap_to_plot          = "mde_incremental",
  cells_to_highlight    = cells_is_sfb,
  highlight_column_name = "annotation_level2",
  labelclusters = FALSE,
  mycols = palette_vector,
  highlight_size = 1.0,
  highlight_alpha = 1.0,
  background_alpha = 1.0,
  pixels = c(512,512)
)
p <- print(plots_is_sfb$plot1 + NoAxes() + NoGrid()) + NoLegend() + theme(legend.position = "none") + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5))

print(p)

#np ova 7dpi
cd4_object$is_npova <- cd4_object$condition_detailed %in% c("NP_OVA_D7")

cells_npova <- colnames(cd4_object)[cd4_object$is_npova]
plots_npova <- MyDimPlotHighlight(
  seurat_object         = cd4_object,
  umap_to_plot          = "mde_incremental",
  cells_to_highlight    = cells_npova,
  highlight_column_name = "annotation_level2",
  labelclusters = FALSE,
  mycols = palette_vector,
  highlight_size = 0.5,
  highlight_alpha = 0.8,
  background_alpha = 1.0,
  pixels = c(512,512)
)
p <- print(plots_npova$plot1 + NoAxes() + NoGrid()) + NoLegend() + theme(legend.position = "none") + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5))

print(p)
```
### Fig 3e
```{r 3e - effector gene/cytokine heatmap, fig.width=8, fig.height=4.5}
sub <- subset(cd4_object, subset = annotation_level2 %in% c("CD4.G", "CD4.H", "CD4.I", "CD4.Q", "CD4.R", "CD4.S", "CD4.T", "CD4.U", "CD4.X", "CD4.Y"))

genes <- c("Ifng","Il4","Il5","Il17a","Il17f","Gzma","Gzmb","Il4", "Il5","Il10","Il21","Il22", "Il13")
gene_order <- c("Il21", "Ifng", "Gzma", "Gzmb","Il10" ,"Il4", "Il5", "Il13", "Il17a", "Il17f", "Il22")
cluster_order <- c("CD4.G","CD4.H","CD4.I","CD4.Q","CD4.R", "CD4.S","CD4.T","CD4.U","CD4.X","CD4.Y")

effector <- FetchData(sub, vars = c(genes, "annotation_level2"))
effector <- effector %>%
  filter(annotation_level2 %in% cluster_order)
eff_long <- effector %>%
  pivot_longer(
    cols = all_of(genes),
    names_to = "gene",
    values_to = "expr"
  )
eff_long <- eff_long %>%
  mutate(expr_bin = expr > 0)

counts <- eff_long %>%
  filter(expr_bin) %>%
  group_by(gene, annotation_level2) %>%
  summarise(n = n(), .groups = "drop")

props <- counts %>%
  group_by(gene) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

props$annotation_level2 <- factor(props$annotation_level2,
                                  levels = cluster_order)

props <- props %>%
  mutate(
    annotation_level2 = factor(annotation_level2, levels = cluster_order),
    gene = factor(gene, levels = gene_order)
  ) %>%
  droplevels()


props_full <- props %>%
  mutate(
    annotation_level2 = factor(annotation_level2, levels = cluster_order),
    gene = factor(gene, levels = gene_order)
  ) %>%
  complete(gene, annotation_level2, fill = list(prop = 0))

mat_pct <- props_full %>%
  mutate(pct = prop * 100) %>%
  select(gene, annotation_level2, pct) %>%
  pivot_wider(names_from = annotation_level2, values_from = pct) %>%
  arrange(gene)

mat <- as.matrix(mat_pct[,-1])
rownames(mat) <- as.character(mat_pct$gene)

white_red <- colorRampPalette(c("white", "red"))(101)

display_numbers <- matrix("", nrow = nrow(mat), ncol = ncol(mat))

display_numbers[mat > 0 & mat < 1]  <- sprintf("%.2f", mat[mat > 0 & mat < 1])
display_numbers[mat > 1 & mat < 10]  <- sprintf("%.1f", mat[mat > 1 & mat < 10])
display_numbers[mat >= 10]           <- sprintf("%.0f", mat[mat >= 10])

display_numbers <- matrix(display_numbers, nrow = nrow(mat)) 

pheatmap(
  mat,
  color = white_red,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  border_color = "darkgray",
  display_numbers = display_numbers,
  number_color = "black",
  fontsize_number = 7,
  angle_col = 45,
  cellwidth = 20,
  cellheight = 20
)

```
### Fig 3f
```{r 3f boxplots showing props of polarized clusters across organs, fig.width=8, fig.height=5}
ann_col   <- "annotation_level2"
sample_col <- "sample_code"
organ_col  <- "organ_simplified"
cond_col   <- "condition_broad"
activated     <- c("CD4.G","CD4.H","CD4.I","CD4.R","CD4.S","CD4.T","CD4.U","CD4.X","CD4.Y")
polarized <- c("CD4.R","CD4.S","CD4.T","CD4.U","CD4.X","CD4.Y", "CD4.H", "CD4.I")

organ_order <- c(
  "spleen","LN","blood","bone marrow","thymus","lung",
  "colon LP","small intestine LP","skin","kidney","liver",
  "pancreas","submandibular gland","synovial fluid","placenta",
  "uterus","prostate","mammary gland","peritoneal cavity","CNS"
)

props2 <- FetchData(cd4_object, vars = c(ann_col, sample_col, organ_col, cond_col)) %>%
  rownames_to_column("cell") %>%
  transmute(
    cell,
    annotation  = .data[[ann_col]],
    sample_code = .data[[sample_col]],
    organ       = .data[[organ_col]],
    condition   = .data[[cond_col]]
  ) %>%
  filter(
    !is.na(sample_code), !is.na(organ), !is.na(condition),
    condition == "healthy",
    annotation %in% activated
  ) %>%
  group_by(organ, sample_code) %>%
  summarise(
    prop = sum(annotation %in% polarized) / n(),
    .groups = "drop"
  ) %>%
  mutate(organ = factor(organ, levels = organ_order)) %>%
  filter(organ %in% organ_order)

p <- ggplot(props2, aes(x = organ, y = prop, fill = organ)) +
  geom_boxplot(width = 0.6, outlier.shape = NA, alpha = 0.8) +
  geom_point(position = position_jitter(width = 0.15), size = 0.6) +
  scale_fill_manual(values = palette_vector_organ) +
  scale_y_continuous(limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  theme_classic() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(x = NULL, y = "% polarized within activated CD4 T Cells (healthy)")

print(p)

```
### Fig 3g
```{r 3g - boxplots showing prop of polarized across activated cells in healthy vs nonhealthy colon, lung, spleen}
ann_col    <- "annotation_level2"
sample_col <- "sample_code"
organ_col  <- "organ_simplified"
cond_col   <- "condition_broad"

activated     <- c("CD4.G","CD4.H","CD4.I","CD4.R","CD4.S","CD4.T","CD4.U","CD4.X","CD4.Y")
polarized <- c("CD4.R","CD4.S","CD4.T","CD4.U","CD4.X","CD4.Y")

organs_keep  <- c("spleen", "colon LP", "lung")
cond_levels  <- c("healthy", "nonhealthy")

df_all <- FetchData(
  cd4_object,
  vars = c(ann_col, sample_col, organ_col, cond_col),
  layer = "data"
) %>%
  rownames_to_column("cell") %>%
  transmute(
    cell,
    annotation  = .data[[ann_col]],
    sample_code = .data[[sample_col]],
    organ       = .data[[organ_col]],
    condition   = .data[[cond_col]]
  ) %>%
  filter(
    !is.na(sample_code),
    !is.na(organ),
    !is.na(condition),
    organ %in% organs_keep,
    annotation %in% activated
  ) %>%
  mutate(
    cond2 = if_else(condition == "healthy", "healthy", "nonhealthy")
  )

props2 <- df_all %>%
  group_by(organ, sample_code, cond2) %>%
  summarise(
    n_activated = n(),
    n_polarized  = sum(annotation %in% polarized),
    prop         = n_polarized / n_activated,
    .groups = "drop"
  ) %>%
  mutate(
    organ = factor(organ, levels = organs_keep),
    cond2 = factor(cond2, levels = cond_levels),
    organ_cond = factor(
      paste(organ, cond2),
      levels = as.vector(t(outer(organs_keep, cond_levels, paste)))
    )
  )

p <- ggplot(props2, aes(x = organ_cond, y = prop, fill = organ)) +
  geom_boxplot(width = 0.6, outlier.shape = NA, alpha = 0.8) +
  geom_point(
    position = position_jitter(width = 0.15),
    size  = 0.6,
    alpha = 1.0
  ) +
  scale_fill_manual(values = palette_vector_organ) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 60, hjust = 1)
  ) +
  labs(
    x = NULL,
    y = "% polarized within activated CD4 T Cells (healthy vs nonhealthy)"
  )

print(p)

```
### Fig 3h
```{r 3h - david}

```
### Fig 3i-k
```{r 3i-k - odhran/christophe}

```
# Figure 4

## T helper programs vary in intensity and combination across mixed and polarized clusters

### Fig 4a-d

```{r 4a-d - polarized cluster signature volcanos - david}
```

### Fig 4e
```{r 4e - single cell heatmap of programs across clusters}
cluster_order <- c(
  "CD4.R","CD4.S","CD4.T","CD4.U","CD4.X",
  "CD4.Y","CD4.H","CD4.I","CD4.Q","CD4.G"
)

score_cols <- c(
  "TH1_Score1",
  "TH2_Score1",
  "TH17_Score1",
  "TFH_TCHRON_Score1"
)

q_no_module_thresh <- 0
g_no_module_thresh <- 0


activated <- subset(
  cd4_object,
  subset = annotation_level2 %in% cluster_order
)

activated$annotation_level2 <- factor(
  activated$annotation_level2,
  levels = cluster_order
)

Idents(activated) <- activated$annotation_level2


# downsample to max 10k cells per cluster
set.seed(123)

cells_keep <- unlist(
  lapply(cluster_order, function(cl) {

    cl_cells <- WhichCells(
      activated,
      idents = cl
    )

    if (length(cl_cells) > 10000) {
      sample(cl_cells, 10000)
    } else {
      cl_cells
    }

  })
)

sub <- subset(
  activated,
  cells = cells_keep
)

sub$annotation_level2 <- factor(
  sub$annotation_level2,
  levels = cluster_order
)

Idents(sub) <- sub$annotation_level2


# make module-score assay from scores already stored in metadata
score_mat <- t(
  as.matrix(
    sub@meta.data[, score_cols]
  )
)

colnames(score_mat) <- colnames(sub)

rownames(score_mat) <- c(
  "Th1 program",
  "Th2 program",
  "Th17 program",
  "Tfh/Tchron program"
)

sub[["ModuleScore"]] <- CreateAssayObject(
  data = score_mat
)

DefaultAssay(sub) <- "ModuleScore"

sub <- ScaleData(
  sub,
  features = rownames(sub[["ModuleScore"]]),
  do.center = TRUE,
  do.scale = TRUE
)


# polyfunctional annotations
prog_df <- polyfunctional_meta %>%
  dplyr::select(
    cell,
    annotation_level2_program
  ) %>%
  distinct(
    cell,
    .keep_all = TRUE
  ) %>%
  mutate(
    annotation_level2_program_clean =
      tolower(annotation_level2_program)
  )


# remove these first if they happen to already exist in metadata
meta <- sub@meta.data %>%
  rownames_to_column("cell") %>%
  dplyr::select(
    -any_of(
      c(
        "annotation_level2_program",
        "annotation_level2_program_clean"
      )
    )
  ) %>%
  left_join(
    prog_df,
    by = "cell"
  ) %>%
  mutate(

    q_max_program = pmax(
      TH1_Score1,
      TH2_Score1,
      TH17_Score1,
      TFH_TCHRON_Score1,
      na.rm = TRUE
    ),

    g_max_program = pmax(
      TH1_Score1,
      TH2_Score1,
      TH17_Score1,
      TFH_TCHRON_Score1,
      na.rm = TRUE
    ),

    is_q_single_or_none =
      annotation_level2_program_clean %in%
      c(
        "cd4.q.none",
        "cd4.q.th1",
        "cd4.q.th2",
        "cd4.q.th17",
        "cd4.q.tfhchron",
        "cd4.q.tfh_tchron"
      ),

    is_g_single_or_none =
      annotation_level2_program_clean %in%
      c(
        "cd4.g.none",
        "cd4.g.th1",
        "cd4.g.th2",
        "cd4.g.th17",
        "cd4.g.tfhchron",
        "cd4.g.tfh_tchron"
      ),

    is_q_polyfunctional =
      annotation_level2 == "CD4.Q" &
      !is.na(annotation_level2_program_clean) &
      grepl(
        "^cd4\\.q\\.",
        annotation_level2_program_clean
      ) &
      !is_q_single_or_none,

    is_g_polyfunctional =
      annotation_level2 == "CD4.G" &
      !is.na(annotation_level2_program_clean) &
      grepl(
        "^cd4\\.g\\.",
        annotation_level2_program_clean
      ) &
      !is_g_single_or_none,

    q_program = case_when(

      annotation_level2 != "CD4.Q" ~
        NA_character_,

      is_q_polyfunctional ~
        "CD4.Q_polyfunctional",

      q_max_program < q_no_module_thresh ~
        "CD4.Q_none",

      TH1_Score1 >= TH2_Score1 &
        TH1_Score1 >= TH17_Score1 &
        TH1_Score1 >= TFH_TCHRON_Score1 ~
        "CD4.Q_th1",

      TH2_Score1 >= TH1_Score1 &
        TH2_Score1 >= TH17_Score1 &
        TH2_Score1 >= TFH_TCHRON_Score1 ~
        "CD4.Q_th2",

      TH17_Score1 >= TH1_Score1 &
        TH17_Score1 >= TH2_Score1 &
        TH17_Score1 >= TFH_TCHRON_Score1 ~
        "CD4.Q_th17",

      TFH_TCHRON_Score1 >= TH1_Score1 &
        TFH_TCHRON_Score1 >= TH2_Score1 &
        TFH_TCHRON_Score1 >= TH17_Score1 ~
        "CD4.Q_tfh_tchron",

      TRUE ~
        "CD4.Q_none"
    ),

    g_program = case_when(

      annotation_level2 != "CD4.G" ~
        NA_character_,

      is_g_polyfunctional ~
        "CD4.G_polyfunctional",

      g_max_program < g_no_module_thresh ~
        "CD4.G_none",

      TH1_Score1 >= TH2_Score1 &
        TH1_Score1 >= TH17_Score1 &
        TH1_Score1 >= TFH_TCHRON_Score1 ~
        "CD4.G_th1",

      TH2_Score1 >= TH1_Score1 &
        TH2_Score1 >= TH17_Score1 &
        TH2_Score1 >= TFH_TCHRON_Score1 ~
        "CD4.G_th2",

      TH17_Score1 >= TH1_Score1 &
        TH17_Score1 >= TH2_Score1 &
        TH17_Score1 >= TFH_TCHRON_Score1 ~
        "CD4.G_th17",

      TFH_TCHRON_Score1 >= TH1_Score1 &
        TFH_TCHRON_Score1 >= TH2_Score1 &
        TFH_TCHRON_Score1 >= TH17_Score1 ~
        "CD4.G_tfh_tchron",

      TRUE ~
        "CD4.G_none"
    ),

    annotation_heatmap = case_when(

      annotation_level2 %in%
        c(
          "CD4.R","CD4.S",
          "CD4.T","CD4.U",
          "CD4.X","CD4.Y",
          "CD4.H","CD4.I"
        ) ~
        as.character(annotation_level2),

      annotation_level2 == "CD4.Q" ~
        q_program,

      annotation_level2 == "CD4.G" ~
        g_program,

      TRUE ~
        as.character(annotation_level2)
    )
  )


rownames(meta) <- meta$cell
meta$cell <- NULL

sub@meta.data <- meta


heatmap_order <- c(
  "CD4.R",
  "CD4.S",
  "CD4.T",
  "CD4.U",
  "CD4.X",
  "CD4.Y",
  "CD4.H",
  "CD4.I",
  "CD4.Q_none",
  "CD4.Q_th1",
  "CD4.Q_th2",
  "CD4.Q_th17",
  "CD4.Q_tfh_tchron",
  "CD4.Q_polyfunctional",
  "CD4.G_none",
  "CD4.G_th1",
  "CD4.G_th2",
  "CD4.G_th17",
  "CD4.G_tfh_tchron",
  "CD4.G_polyfunctional"
)

sub$annotation_heatmap <- factor(
  sub$annotation_heatmap,
  levels = heatmap_order
)


meta <- sub@meta.data %>%
  mutate(
    cell = rownames(.)
  )

set.seed(123)

randomize_per_group <- function(
  df,
  group_name
) {

  cells_gr <- df %>%
    filter(
      annotation_heatmap == group_name
    ) %>%
    pull(cell)

  if (length(cells_gr) == 0) {
    return(NULL)
  }

  sample(cells_gr)
}


cell_order <- c(
  randomize_per_group(meta, "CD4.R"),
  randomize_per_group(meta, "CD4.S"),
  randomize_per_group(meta, "CD4.T"),
  randomize_per_group(meta, "CD4.U"),
  randomize_per_group(meta, "CD4.X"),
  randomize_per_group(meta, "CD4.Y"),
  randomize_per_group(meta, "CD4.H"),
  randomize_per_group(meta, "CD4.I"),
  randomize_per_group(meta, "CD4.Q_none"),
  randomize_per_group(meta, "CD4.Q_th1"),
  randomize_per_group(meta, "CD4.Q_th2"),
  randomize_per_group(meta, "CD4.Q_th17"),
  randomize_per_group(meta, "CD4.Q_tfh_tchron"),
  randomize_per_group(meta, "CD4.Q_polyfunctional"),
  randomize_per_group(meta, "CD4.G_none"),
  randomize_per_group(meta, "CD4.G_th1"),
  randomize_per_group(meta, "CD4.G_th2"),
  randomize_per_group(meta, "CD4.G_th17"),
  randomize_per_group(meta, "CD4.G_tfh_tchron"),
  randomize_per_group(meta, "CD4.G_polyfunctional")
)

cell_order <- unique(cell_order)

cell_order <- cell_order[
  cell_order %in% colnames(sub)
]


module_features <- rownames(
  sub[["ModuleScore"]]
)


DoHeatmap(
  sub,
  features = module_features,
  group.by = "annotation_heatmap",
  assay = "ModuleScore",
  cells = cell_order,
  disp.min = -2,
  disp.max = 2,
  raster = FALSE,
  angle = 90
) +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  NoLegend()
```
### Fig 4f-i
```{r 4f-i scatter plots of module scores - all cluster means, only plot g/q cells}
clusters_keep <- c("CD4.Q","CD4.G")

modulescores <- FetchData(
  cd4_object,
  vars = c("TH1_Score1", "TH2_Score1", "TH17_Score1", "TFH_TCHRON_Score1", "annotation_level2")
) %>%
  filter(annotation_level2 %in% clusters_keep)

modulescores <- modulescores %>%
  mutate(
    TH1_Score1 = pmin(TH1_Score1, 2.0),
    TH2_Score1 = pmin(TH2_Score1, 2.0),
    TH17_Score1 = pmin(TH17_Score1, 2.0),
    TFH_TCHRON_Score1 = pmin(TFH_TCHRON_Score1, 2.0)
  )

mean <- modulescores %>%
  group_by(annotation_level2) %>%
  summarise(
    TH1 = mean(TH1_Score1, na.rm = TRUE),
    TH2 = mean(TH2_Score1, na.rm = TRUE),
    TH17 = mean(TH17_Score1, na.rm = TRUE),
    TFH_TCHRON = mean(TFH_TCHRON_Score1, na.rm = TRUE)
  )

clusters_keep_2 <- c("CD4.H","CD4.I", "CD4.R", "CD4.S", "CD4.T", "CD4.U", "CD4.X", "CD4.Y")

modulescores2 <- FetchData(
  cd4_object,
  vars = c("TH1_Score1", "TH2_Score1", "TH17_Score1", "TFH_TCHRON_Score1", "annotation_level2")
) %>%
  filter(annotation_level2 %in% clusters_keep_2)


mean2 <- modulescores2 %>%
  group_by(annotation_level2) %>%
  summarise(
    TH1 = mean(TH1_Score1, na.rm = TRUE),
    TH2 = mean(TH2_Score1, na.rm = TRUE),
    TH17 = mean(TH17_Score1, na.rm = TRUE),
    TFH_TCHRON = mean(TFH_TCHRON_Score1, na.rm = TRUE)
  )


quadrant_pct <- modulescores %>%
  mutate(quadrant = case_when(
    TH2_Score1 >= 0.25 & TH1_Score1 >= 0 ~ "Q1 (TH2+TH1+)",
    TH2_Score1 <  0.25 & TH1_Score1 >= 0 ~ "Q2 (TH2-TH1+)",
    TH2_Score1 <  0.25 & TH1_Score1 <  0 ~ "Q3 (TH2-TH1-)",
    TH2_Score1 >= 0.25 & TH1_Score1 <  0 ~ "Q4 (TH2+TH1-)"
  )) %>%
  count(quadrant) %>%
  mutate(pct = n / sum(n) * 100,
         label = sprintf("%.1f%%", pct))

quadrant_coords <- data.frame(
  quadrant = c("Q1 (TH2+TH1+)", "Q2 (TH2-TH1+)", "Q3 (TH2-TH1-)", "Q4 (TH2+TH1-)"),
  x = c(1.2,  -0.1,  -0.1,  1.2),
  y = c(1.0,   1.3,  -0.2,  -0.2)
)

quadrant_pct <- left_join(quadrant_pct, quadrant_coords, by = "quadrant")

ggplot(modulescores, aes(x = TH2_Score1, y = TH1_Score1, color = annotation_level2)) +
  geom_point_rast(size = 0.4, alpha = 0.3, raster.dpi = 200) + 
  geom_point(
    data = mean,
    aes(x = TH2, y = TH1, fill = annotation_level2),
    shape = 21, color = "black", size = 4, stroke = 1
  ) + 
  geom_point(
    data = mean2,
    aes(x = TH2, y = TH1, fill = annotation_level2),
    shape = 21, color = "black", size = 4, stroke = 1
  ) +
  geom_text(
    data = mean,
    aes(x = TH2, y = TH1, label = annotation_level2),
    color = "black", vjust = -1, size = 4
  ) +
  geom_text(
    data = mean2,
    aes(x = TH2, y = TH1, label = annotation_level2),
    color = "black", vjust = -1, size = 4
  ) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", linewidth = 0.4) +
  geom_vline(xintercept = 0.25, color = "red", linetype = "dashed", linewidth = 0.4) +
  geom_text(
    data = quadrant_pct,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    color = "black", fontface = "bold", size = 4
  ) +
  theme_classic() +
  labs(x = "Th2-like module score", y = "Th1-like module score",
       color = "Cluster", fill = "Cluster") +
  coord_cartesian(xlim = c(-0.2, 1.5), ylim = c(-0.5, 1.5)) +
  scale_color_manual(values = palette_vector) +
  scale_fill_manual(values = palette_vector)

#th1 vs th17
quadrant_pct <- modulescores %>%
  mutate(quadrant = case_when(
    TH17_Score1 >= 0.25 & TH1_Score1 >= 0 ~ "Q1 (TH17+TH1+)",
    TH17_Score1 <  0.25 & TH1_Score1 >= 0 ~ "Q2 (TH17-TH1+)",
    TH17_Score1 <  0.25 & TH1_Score1 <  0 ~ "Q3 (TH17-TH1-)",
    TH17_Score1 >= 0.25 & TH1_Score1 <  0 ~ "Q4 (TH17+TH1-)"
  )) %>%
  count(quadrant) %>%
  mutate(pct = n / sum(n) * 100,
         label = sprintf("%.1f%%", pct))

quadrant_coords <- data.frame(
  quadrant = c("Q1 (TH17+TH1+)", "Q2 (TH17-TH1+)", "Q3 (TH17-TH1-)", "Q4 (TH17+TH1-)"),
  x = c(1.2,  -0.1,  -0.1,  1.2),
  y = c(1.0,   1.3,  -0.2,  -0.2)
)


quadrant_pct <- left_join(quadrant_pct, quadrant_coords, by = "quadrant")

ggplot(modulescores, aes(x = TH17_Score1, y = TH1_Score1, color = annotation_level2)) +
  geom_point_rast(size = 0.4, alpha = 0.3, raster.dpi = 200) + 
  geom_point(
    data = mean,
    aes(x = TH17, y = TH1, fill = annotation_level2),
    shape = 21, color = "black", size = 4, stroke = 1
  )  + 
  geom_point(
    data = mean2,
    aes(x = TH17, y = TH1, fill = annotation_level2),
    shape = 21, color = "black", size = 4, stroke = 1
  ) +
  geom_text(
    data = mean,
    aes(x = TH17, y = TH1, label = annotation_level2),
    color = "black", vjust = -1, size = 4
  ) +
  geom_text(
    data = mean2,
    aes(x = TH17, y = TH1, label = annotation_level2),
    color = "black", vjust = -1, size = 4
  ) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", linewidth = 0.4) +
  geom_vline(xintercept = 0.25, color = "red", linetype = "dashed", linewidth = 0.4) +
  
  geom_text(
    data = quadrant_pct,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    color = "black", fontface = "bold", size = 4
  ) +
  theme_classic() +
  labs(x = "Th17-like module score", y = "Th1-like module score",
       color = "Cluster", fill = "Cluster") +
  coord_cartesian(xlim = c(-0.3, 1.5), ylim = c(-0.5, 1.5)) +
  scale_color_manual(values = palette_vector) +
  scale_fill_manual(values = palette_vector)

#th2 vs th17
quadrant_pct <- modulescores %>%
  mutate(quadrant = case_when(
    TH17_Score1 >= 0.25 & TH2_Score1 >= 0.25 ~ "Q1 (TH17+TH2+)",
    TH17_Score1 <  0.25 & TH2_Score1 >= 0.25 ~ "Q2 (TH17-TH2+)",
    TH17_Score1 <  0.25 & TH2_Score1 <  0.25 ~ "Q3 (TH17-TH2-)",
    TH17_Score1 >= 0.25 & TH2_Score1 <  0.25 ~ "Q4 (TH17+TH2-)"
  )) %>%
  count(quadrant) %>%
  mutate(pct = n / sum(n) * 100,
         label = sprintf("%.1f%%", pct))

quadrant_coords <- data.frame(
  quadrant = c("Q1 (TH17+TH2+)", "Q2 (TH17-TH2+)", "Q3 (TH17-TH2-)", "Q4 (TH17+TH2-)"),
  x = c(1.2,  -0.1,  -0.1,  1.2),
  y = c(1.0,   1.3,  -0.2,  -0.2)
)


quadrant_pct <- left_join(quadrant_pct, quadrant_coords, by = "quadrant")

ggplot(modulescores, aes(x = TH17_Score1, y = TH2_Score1, color = annotation_level2)) +
  geom_point_rast(size = 0.4, alpha = 0.3, raster.dpi = 200) + 
  geom_point(
    data = mean,
    aes(x = TH17, y = TH2, fill = annotation_level2),
    shape = 21, color = "black", size = 4, stroke = 1
  )  + 
  geom_point(
    data = mean2,
    aes(x = TH17, y = TH2, fill = annotation_level2),
    shape = 21, color = "black", size = 4, stroke = 1
  ) +
  geom_text(
    data = mean,
    aes(x = TH17, y = TH2, label = annotation_level2),
    color = "black", vjust = -1, size = 4
  ) +
  geom_text(
    data = mean2,
    aes(x = TH17, y = TH2, label = annotation_level2),
    color = "black", vjust = -1, size = 4
  ) +
  geom_hline(yintercept = 0.25, color = "red", linetype = "dashed", linewidth = 0.4) +
  geom_vline(xintercept = 0.25, color = "red", linetype = "dashed", linewidth = 0.4) +
  
  geom_text(
    data = quadrant_pct,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    color = "black", fontface = "bold", size = 4
  ) +
  theme_classic() +
  labs(x = "Th17-like module score", y = "Th2-like module score",
       color = "Cluster", fill = "Cluster") +
  coord_cartesian(xlim = c(-0.3, 1.5), ylim = c(-0.2, 1.5)) +
  scale_color_manual(values = palette_vector) +
  scale_fill_manual(values = palette_vector)

#th1 vs tfhchron
quadrant_pct <- modulescores %>%
  mutate(quadrant = case_when(
    TFH_TCHRON_Score1 >= 0.25 & TH1_Score1 >= 0 ~ "Q1 (TFH_TCHRON+TH1+)",
    TFH_TCHRON_Score1 <  0.25 & TH1_Score1 >= 0 ~ "Q2 (TFH_TCHRON-TH1+)",
    TFH_TCHRON_Score1 <  0.25 & TH1_Score1 <  0 ~ "Q3 (TFH_TCHRON-TH1-)",
    TFH_TCHRON_Score1 >= 0.25 & TH1_Score1 <  0 ~ "Q4 (TFH_TCHRON+TH1-)"
  )) %>%
  count(quadrant) %>%
  mutate(pct = n / sum(n) * 100,
         label = sprintf("%.1f%%", pct))

quadrant_coords <- data.frame(
  quadrant = c("Q1 (TFH_TCHRON+TH1+)", "Q2 (TFH_TCHRON-TH1+)", "Q3 (TFH_TCHRON-TH1-)", "Q4 (TFH_TCHRON+TH1-)"),
  x = c(1.2,  -0.1,  -0.1,  1.2),
  y = c(1.0,   1.3,  -0.2,  -0.2)
)

quadrant_pct <- left_join(quadrant_pct, quadrant_coords, by = "quadrant")

ggplot(modulescores, aes(x = TFH_TCHRON_Score1, y = TH1_Score1, color = annotation_level2)) +
  geom_point_rast(size = 0.4, alpha = 0.3, raster.dpi = 200) + 
  geom_point(
    data = mean,
    aes(x = TFH_TCHRON, y = TH1, fill = annotation_level2),
    shape = 21, color = "black", size = 4, stroke = 1
  )  + 
  geom_point(
    data = mean2,
    aes(x = TFH_TCHRON, y = TH1, fill = annotation_level2),
    shape = 21, color = "black", size = 4, stroke = 1
  ) +
  geom_text(
    data = mean,
    aes(x = TFH_TCHRON, y = TH1, label = annotation_level2),
    color = "black", vjust = -1, size = 4
  ) +
  geom_text(
    data = mean2,
    aes(x = TFH_TCHRON, y = TH1, label = annotation_level2),
    color = "black", vjust = -1, size = 4
  ) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", linewidth = 0.4) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", linewidth = 0.4) +
  
  geom_text(
    data = quadrant_pct,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    color = "black", fontface = "bold", size = 4
  ) +
  theme_classic() +
  labs(x = "Tfh/Tchron-like module score", y = "Th1-like module score",
       color = "Cluster", fill = "Cluster") +
  coord_cartesian(xlim = c(-0.50, 1.5), ylim = c(-0.5, 1.5)) +
  scale_color_manual(values = palette_vector) +
  scale_fill_manual(values = palette_vector)
```
### Fig 4j-m
```{r 4j-m David}

```
# Figure 5

## Convergence of Tfh and chronically stimulated CD4+ T cells into CD4.H and CD4.I

### Fig 5a-f
```{r 5a-f - Julia}

```
### Fig 5g
```{r 5g feature scatter plots of pd1 x cxcr5 protein expression in cd4.h/i, fig.width = 6, fig.height=6}
cd4_cite <- NormalizeData(cd4_cite, assay = "ADT", normalization.method = "LogNormalize")

cd4_cite$h <- cd4_cite$annotation_level2 == "CD4.H"

p <- MyFeatureScatter(
  so = cd4_cite,
  group.by = "h",
  feature1 = "CXCR5",
  feature2 = "PDCD1.PD1.CD279",
  raster = TRUE
) + coord_cartesian(xlim = c(0, 5))

print(p)


cd4_cite$i <- cd4_cite$annotation_level2  == "CD4.I"

p <- MyFeatureScatter(
  so = cd4_cite,
  group.by = "i",
  feature1 = "CXCR5",
  feature2 = "PDCD1.PD1.CD279",
  raster = TRUE
) + coord_cartesian(xlim = c(0, 5))

print(p)
```
### Fig 5h
```{r 5h feature scatter plots of fr4 x cd73 protein expression in cd4.h/i, fig.width = 6, fig.height= 6}
p <- MyFeatureScatter(
  so = cd4_cite,
  group.by = "i",
  feature1 = "FR4",
  feature2 = "CD73.5NTD",
  raster = TRUE
)
print(p)

p <- MyFeatureScatter(
  so = cd4_cite,
  group.by = "h",
  feature1 = "FR4",
  feature2 = "CD73.5NTD",
  raster = TRUE
)
print(p)
```
# Figure 6

## Memory CD4+ T cells map to the same clusters as effector cells

### Fig 6a
```{r 6a - mdes mapping time course of slo listeria Tet+ Cells, fig.width=8, fig.height=8}
conds <- cd4_object@meta.data %>%
  filter(igt_number == "43") %>%
  pull(condition_detailed) %>%
  unique() %>%
  sort()

for (cond in conds) {
  
  message("Plotting: ", cond)
  
  cells_cond <- colnames(cd4_object)[
    cd4_object$igt_number == "43" &
    cd4_object$condition_detailed == cond
  ]
  
  plots_cond <- MyDimPlotHighlight(
    seurat_object         = cd4_object,
    umap_to_plot          = "mde_incremental",
    cells_to_highlight    = cells_cond,
    highlight_column_name = "annotation_level2",
    title                 = cond,
    labelclusters         = FALSE,
    highlight_raster      = TRUE,
    highlight_pixels      = c(1024, 1024),
    mycols                = palette_vector,
    highlight_size        = 4,
    highlight_alpha       = 1,
    background_alpha      = 1.0,
    pixels                = c(512, 512)
  )
  
  p <- plots_cond$plot1 +
    NoAxes() +
    NoLegend() +
    theme(legend.position = "none") +
    coord_cartesian(
      xlim = c(-2.5, 4),
      ylim = c(-2.5, 2.5)
    )
  

  
  print(p)
  
}
```
### Fig 6c
```{r 6c - acute listeria module score scatter plots}
listeria_object <- subset(cd4_object, subset = igt_number == "43")
listeria_ids <- listeria_ids <- colnames(listeria_object)

listeria_meta <- listeria_object@meta.data %>%
  rownames_to_column("cell")

clusters_keep <- c("CD4.Q","CD4.G","CD4.H", "CD4.I", "CD4.R", "CD4.S", "CD4.T", "CD4.U", "CD4.X", "CD4.Y")

modulescores <- FetchData(
  cd4_object,
  vars = c(
    "TH1_Score1",
    "TH2_Score1",
    "TH17_Score1",
    "TFH_TCHRON_Score1",
    "annotation_level2",
    "organ_simplified",
    "condition_detailed"
  )
) %>%
  rownames_to_column("cell") %>%
  mutate(is_listeria = cell %in% listeria_ids) %>%
  filter(annotation_level2 %in% clusters_keep)


listeria_d7_slo <- modulescores %>%
  filter(
    is_listeria,
    organ_simplified %in% c("spleen", "SLO"),
    condition_detailed %in% c("Lmonocytogenes_D6", "Lmonocytogenes_D10")
  )

quadrant_pct <- listeria_d7_slo %>%
  mutate(
    quadrant = case_when(
      TFH_TCHRON_Score1 >= 0 & TH1_Score1 >= 0 ~ "Q1 (TFH_TCHRON+TH1+)",
      TFH_TCHRON_Score1 <  0 & TH1_Score1 >= 0 ~ "Q2 (TFH_TCHRON-TH1+)",
      TFH_TCHRON_Score1 <  0 & TH1_Score1 <  0 ~ "Q3 (TFH_TCHRON-TH1-)",
      TFH_TCHRON_Score1 >= 0 & TH1_Score1 <  0 ~ "Q4 (TFH_TCHRON+TH1-)"
    )
  ) %>%
  count(quadrant) %>%
  mutate(
    pct = n / sum(n) * 100,
    label = sprintf("%.1f%%", pct)
  )

quadrant_coords <- data.frame(
  quadrant = c("Q1 (TFH_TCHRON+TH1+)", "Q2 (TFH_TCHRON-TH1+)", "Q3 (TFH_TCHRON-TH1-)", "Q4 (TFH_TCHRON+TH1-)"),
  x = c(1.2, -0.1, -0.1, 1.2),
  y = c(1.0, 1.3, -0.2, -0.2)
)

quadrant_pct <- left_join(quadrant_pct, quadrant_coords, by = "quadrant")

mean_listeria <- listeria_d7_slo %>%
  group_by(annotation_level2) %>%
  summarise(
    TFH_TCHRON = mean(TFH_TCHRON_Score1, na.rm = TRUE),
    TH1 = mean(TH1_Score1, na.rm = TRUE),
    .groups = "drop"
  )

ggplot() +
  geom_point_rast(
    data = modulescores,
    aes(x = TFH_TCHRON_Score1, y = TH1_Score1),
    color = "grey80",
    size = 0.35,
    alpha = 0.25,
    raster.dpi = 200
  ) +
  geom_point_rast(
    data = listeria_d7_slo,
    aes(x = TFH_TCHRON_Score1, y = TH1_Score1, color = annotation_level2),
    size = 3,
    alpha = 0.9,
    raster.dpi = 200
  ) +
  geom_point(
    data = mean_listeria,
    aes(x = TFH_TCHRON, y = TH1, fill = annotation_level2),
    shape = 21,
    color = "black",
    size = 4,
    stroke = 1
  ) +
  geom_text(
    data = mean_listeria,
    aes(x = TFH_TCHRON, y = TH1, label = annotation_level2),
    color = "black",
    vjust = -1,
    size = 4
  ) +
  geom_hline(
    yintercept = 0,
    color = "red",
    linetype = "dashed",
    linewidth = 0.4
  ) +
  geom_vline(
    xintercept = 0,
    color = "red",
    linetype = "dashed",
    linewidth = 0.4
  ) +
  geom_text(
    data = quadrant_pct,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    color = "black",
    fontface = "bold",
    size = 4
  ) +
  theme_classic() +
  labs(
    x = "Tfh_Tchron-like module score",
    y = "Th1-like module score",
    color = "Cluster",
    fill = "Cluster"
  ) +
  coord_cartesian(xlim = c(-0.4, 1.5), ylim = c(-0.5, 1.5)) +
  scale_color_manual(values = palette_vector) +
  scale_fill_manual(values = palette_vector)


#th2 vs th17
quadrant_pct <- listeria_d7_slo %>%
  mutate(
    quadrant = case_when(
      TH17_Score1 >= 0.25 & TH2_Score1 >= 0.25 ~ "Q1 (TH17+TH2+)",
      TH17_Score1 <  0.25 & TH2_Score1 >= 0.25 ~ "Q2 (TH17-TH2+)",
      TH17_Score1 <  0.25 & TH2_Score1 <  0.25 ~ "Q3 (TH17-TH2-)",
      TH17_Score1 >= 0.25 & TH2_Score1 <  0.25 ~ "Q4 (TH17+TH2-)"
    )
  ) %>%
  count(quadrant) %>%
  mutate(
    pct = n / sum(n) * 100,
    label = sprintf("%.1f%%", pct)
  )

quadrant_coords <- data.frame(
  quadrant = c("Q1 (TH17+TH2+)", "Q2 (TH17-TH2+)", "Q3 (TH17-TH2-)", "Q4 (TH17+TH2-)"),
  x = c(1.2, -0.1, -0.1, 1.2),
  y = c(1.0, 1.3, -0.2, -0.2)
)

quadrant_pct <- left_join(quadrant_pct, quadrant_coords, by = "quadrant")

mean_listeria <- listeria_d7_slo %>%
  group_by(annotation_level2) %>%
  summarise(
    TH17 = mean(TH17_Score1, na.rm = TRUE),
    TH2 = mean(TH2_Score1, na.rm = TRUE),
    .groups = "drop"
  )


ggplot() +
  geom_point_rast(
    data = modulescores,
    aes(x = TH17_Score1, y = TH2_Score1),
    color = "grey80",
    size = 0.35,
    alpha = 0.25,
    raster.dpi = 200
  ) +
  geom_point_rast(
    data = listeria_d7_slo,
    aes(x = TH17_Score1, y = TH2_Score1, color = annotation_level2),
    size = 3,
    alpha = 0.9,
    raster.dpi = 200
  ) +
  geom_point(
    data = mean_listeria,
    aes(x = TH17, y = TH2, fill = annotation_level2),
    shape = 21,
    color = "black",
    size = 4,
    stroke = 1
  ) +
  geom_text(
    data = mean_listeria,
    aes(x = TH17, y = TH2, label = annotation_level2),
    color = "black",
    vjust = -1,
    size = 4
  ) +
  geom_hline(
    yintercept = 0.25,
    color = "red",
    linetype = "dashed",
    linewidth = 0.4
  ) +
  geom_vline(
    xintercept = 0.25,
    color = "red",
    linetype = "dashed",
    linewidth = 0.4 
  ) +
  geom_text(
    data = quadrant_pct,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    color = "black",
    fontface = "bold",
    size = 4
  ) +
  theme_classic() +
  labs(
    x = "Th17-like module score",
    y = "Th2-like module score",
    color = "Cluster",
    fill = "Cluster"
  ) +
  coord_cartesian(xlim = c(-0.4, 1.5), ylim = c(-0.5, 1.5)) +
  scale_color_manual(values = palette_vector) +
  scale_fill_manual(values = palette_vector)

```
### Fig 6d
```{r 6d memory listeria module score scatter plots}
listeria_25_slo <- modulescores %>%
  filter(
    is_listeria,
    organ_simplified %in% c("spleen", "SLO"),
    condition_detailed %in% c("Lmonocytogenes_D25", "Lmonocytogenes_D30", "Lmonocytogenes_D60")
  )

quadrant_pct <- listeria_25_slo %>%
  mutate(
    quadrant = case_when(
      TFH_TCHRON_Score1 >= 0 & TH1_Score1 >= 0 ~ "Q1 (TFH_TCHRON+TH1+)",
      TFH_TCHRON_Score1 <  0 & TH1_Score1 >= 0 ~ "Q2 (TFH_TCHRON-TH1+)",
      TFH_TCHRON_Score1 <  0 & TH1_Score1 <  0 ~ "Q3 (TFH_TCHRON-TH1-)",
      TFH_TCHRON_Score1 >= 0 & TH1_Score1 <  0 ~ "Q4 (TFH_TCHRON+TH1-)"
    )
  ) %>%
  count(quadrant) %>%
  mutate(
    pct = n / sum(n) * 100,
    label = sprintf("%.1f%%", pct)
  )

quadrant_coords <- data.frame(
  quadrant = c("Q1 (TFH_TCHRON+TH1+)", "Q2 (TFH_TCHRON-TH1+)", "Q3 (TFH_TCHRON-TH1-)", "Q4 (TFH_TCHRON+TH1-)"),
  x = c(1.2, -0.1, -0.1, 1.2),
  y = c(1.0, 1.3, -0.2, -0.2)
)

quadrant_pct <- left_join(quadrant_pct, quadrant_coords, by = "quadrant")

mean_listeria <- listeria_25_slo %>%
  group_by(annotation_level2) %>%
  summarise(
    TFH_TCHRON = mean(TFH_TCHRON_Score1, na.rm = TRUE),
    TH1 = mean(TH1_Score1, na.rm = TRUE),
    .groups = "drop"
  )

ggplot() +
  geom_point_rast(
    data = modulescores,
    aes(x = TFH_TCHRON_Score1, y = TH1_Score1),
    color = "grey80",
    size = 0.35,
    alpha = 0.25,
    raster.dpi = 200
  ) +
  geom_point_rast(
    data = listeria_25_slo,
    aes(x = TFH_TCHRON_Score1, y = TH1_Score1, color = annotation_level2),
    size = 3,
    alpha = 0.9,
    raster.dpi = 200
  ) +
  geom_point(
    data = mean_listeria,
    aes(x = TFH_TCHRON, y = TH1, fill = annotation_level2),
    shape = 21,
    color = "black",
    size = 4,
    stroke = 1
  ) +
  geom_text(
    data = mean_listeria,
    aes(x = TFH_TCHRON, y = TH1, label = annotation_level2),
    color = "black",
    vjust = -1,
    size = 4
  ) +
  geom_hline(
    yintercept = 0,
    color = "red",
    linetype = "dashed",
    linewidth = 0.4
  ) +
  geom_vline(
    xintercept = 0,
    color = "red",
    linetype = "dashed",
    linewidth = 0.4
  ) +
  geom_text(
    data = quadrant_pct,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    color = "black",
    fontface = "bold",
    size = 4
  ) +
  theme_classic() +
  labs(
    x = "Tfh_Tchron-like module score",
    y = "Th1-like module score",
    color = "Cluster",
    fill = "Cluster"
  ) +
  coord_cartesian(xlim = c(-0.4, 1.5), ylim = c(-0.5, 1.5)) +
  scale_color_manual(values = palette_vector) +
  scale_fill_manual(values = palette_vector)


#th2 vs th17
quadrant_pct <- listeria_d7_slo %>%
  mutate(
    quadrant = case_when(
      TH17_Score1 >= 0.25 & TH2_Score1 >= 0.25 ~ "Q1 (TH17+TH2+)",
      TH17_Score1 <  0.25 & TH2_Score1 >= 0.25 ~ "Q2 (TH17-TH2+)",
      TH17_Score1 <  0.25 & TH2_Score1 <  0.25 ~ "Q3 (TH17-TH2-)",
      TH17_Score1 >= 0.25 & TH2_Score1 <  0.25 ~ "Q4 (TH17+TH2-)"
    )
  ) %>%
  count(quadrant) %>%
  mutate(
    pct = n / sum(n) * 100,
    label = sprintf("%.1f%%", pct)
  )

quadrant_coords <- data.frame(
  quadrant = c("Q1 (TH17+TH2+)", "Q2 (TH17-TH2+)", "Q3 (TH17-TH2-)", "Q4 (TH17+TH2-)"),
  x = c(1.2, -0.1, -0.1, 1.2),
  y = c(1.0, 1.3, -0.2, -0.2)
)

quadrant_pct <- left_join(quadrant_pct, quadrant_coords, by = "quadrant")

mean_listeria <- listeria_25_slo %>%
  group_by(annotation_level2) %>%
  summarise(
    TH17 = mean(TH17_Score1, na.rm = TRUE),
    TH2 = mean(TH2_Score1, na.rm = TRUE),
    .groups = "drop"
  )


ggplot() +
  geom_point_rast(
    data = modulescores,
    aes(x = TH17_Score1, y = TH2_Score1),
    color = "grey80",
    size = 0.35,
    alpha = 0.25,
    raster.dpi = 200
  ) +
  geom_point_rast(
    data = listeria_25_slo,
    aes(x = TH17_Score1, y = TH2_Score1, color = annotation_level2),
    size = 3,
    alpha = 0.9,
    raster.dpi = 200
  ) +
  geom_point(
    data = mean_listeria,
    aes(x = TH17, y = TH2, fill = annotation_level2),
    shape = 21,
    color = "black",
    size = 4,
    stroke = 1
  ) +
  geom_text(
    data = mean_listeria,
    aes(x = TH17, y = TH2, label = annotation_level2),
    color = "black",
    vjust = -1,
    size = 4
  ) +
  geom_hline(
    yintercept = 0.25,
    color = "red",
    linetype = "dashed",
    linewidth = 0.4
  ) +
  geom_vline(
    xintercept = 0.25,
    color = "red",
    linetype = "dashed",
    linewidth = 0.4 
  ) +
  geom_text(
    data = quadrant_pct,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    color = "black",
    fontface = "bold",
    size = 4
  ) +
  theme_classic() +
  labs(
    x = "Th17-like module score",
    y = "Th2-like module score",
    color = "Cluster",
    fill = "Cluster"
  ) +
  coord_cartesian(xlim = c(-0.4, 1.5), ylim = c(-0.5, 1.5)) +
  scale_color_manual(values = palette_vector) +
  scale_fill_manual(values = palette_vector)
```
### Fig 6e
```{r 6e mde highlighting cd62l+cd69+ mde, fig.width = 8, fig.height=8}
adt <- GetAssayData(cd4_cite, assay = "ADT", layer = "data")
cd62l_feat <- "CD62L"
cd69_feat <- "CD69"

cd62l_thresh <- 5
cd69_thresh <- 5


cd4_cite$is_cd62l_cd69 <-
  (adt[cd62l_feat, ]  > cd62l_thresh) &
  (adt[cd69_feat, ] > cd69_thresh)

MyDimPlotHighlightDensity(
  seurat_object = cd4_cite,
  umap_to_plot = "mde_incremental",
  group.by = "is_cd62l_cd69",
  split.by = NULL,
  raster = FALSE,
  highlight_size = 0.25,
  highlight_alpha = 1.0,
  cols = magma(100)
)
print(p)
```

# Figure 7

## Resting CD4+ T cell heterogeneity

### Figure 7a
```{r 7a resting cd4 citeseq heatmap, fig.width = 10, fig.height=5}
cd4_cite <- subset(
  cd4_cite,
  subset = !annotation_level2 %in% c("CD4.wN", "CD4.wZ", "CD4.P", "CD4.wM")
)

DefaultAssay(cd4_cite) <- "ADT"

citeseq_markers <- c(
  "CD62L","CD44","FR4","CD55.DAF",
  "LY6C1-LY6C2","CD45RB","CD5","CD69"
)

cluster_order <- levels(factor(cd4_cite$annotation_level2))
cd4_cite$annotation_level2 <- factor(cd4_cite$annotation_level2, levels = cluster_order)
Idents(cd4_cite) <- cd4_cite$annotation_level2

adt_counts <- GetAssayData(cd4_cite, assay = "ADT", layer = "counts")

present_adt <- intersect(citeseq_markers, rownames(adt_counts))
adt_counts <- adt_counts[present_adt, , drop = FALSE]

cell_totals <- Matrix::colSums(GetAssayData(cd4_cite, assay = "ADT", layer = "counts"))
cell_totals[cell_totals == 0] <- NA

adt_cp10k <- sweep(adt_counts, 2, cell_totals, "/") * 10000

avg_cp10k <- sapply(cluster_order, function(clust) {
  cells_use <- WhichCells(cd4_cite, idents = clust)
  if (length(cells_use) == 1) {
    as.numeric(adt_cp10k[, cells_use])
  } else {
    Matrix::rowMeans(adt_cp10k[, cells_use, drop = FALSE], na.rm = TRUE)
  }
})

avg_cp10k <- as.matrix(avg_cp10k)
rownames(avg_cp10k) <- present_adt
colnames(avg_cp10k) <- cluster_order

marker_order <- c(
  "CD62L","CD44","FR4","CD55.DAF",
  "LY6C1-LY6C2","CD45RB","CD5","CD69"
)
marker_order <- marker_order[marker_order %in% rownames(avg_cp10k)]
avg_cp10k <- avg_cp10k[marker_order, , drop = FALSE]
avg_cp10k_log1p <- log1p(avg_cp10k)

blue_white_red <- colorRampPalette(c("blue", "white", "red"))(100)

pheatmap(
  avg_cp10k_log1p,
  color = blue_white_red,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_row = 10,
  fontsize_col = 8,
  angle_col = 45,
  border_color = "darkgray",
  cellwidth = 12,
  cellheight = 12,
  main = "Mean ADT expression (log1p(CP10K))"
)
```
### Fig 7c
```{r 7c iv- clusters in nlts, fig.width = 8, fig.height= 8}
cd4_object$is_resting_ivneg <- cd4_object$annotation_level2 %in% c("CD4.A", "CD4.B", "CD4.C", "CD4.D", "CD4.E", "CD4.F") & cd4_object$gating_strategy %in% c(" IV- CD3+", "IV- CD45.2+ CD3e+", "IV- CD45+ CD3+ CD44+", "IV- CD45+ CD3e+", " IV- CD45+ CD3e+ CD44+ Derp1 Tet +/+", " IV- CD45+ CD3e+ CD44+ Thy1.1- and Thy1.1+ (OT1)", " IV- CD45+ CD3e+ Derp1 tet- and IV- CD45+ CD3e+ Derp1 tet+", "IV- CD45+ CD3e+ Thy1.1- and Thy1.1+ (OT1)")
cells_ivneg <- colnames(cd4_object)[cd4_object$is_resting_ivneg]
plots_ivneg <- MyDimPlotHighlight(
  seurat_object         = cd4_object,
  umap_to_plot          = "mde_incremental",
  cells_to_highlight    = cells_ivneg,
  highlight_column_name = "annotation_level2",
  title                 = "Resting IV- Cells highlighted",
  labelclusters = FALSE,
  highlight_raster = T,
  highlight_pixels =  c(1024, 1024),
  mycols = palette_vector,
  highlight_size = 4,
  highlight_alpha = 1,
  background_alpha = 1.0,
  pixels = c(512,512)
)

p <- print(plots_ivneg$plot1 + NoAxes() + NoGrid()) + NoLegend() + theme(legend.position = "none") + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5))

print(p)
```
### Fig 7d
```{r 7d - top dge between resting clusters}

cluster_order <- c("CD4.A","CD4.B","CD4.C","CD4.D","CD4.E","CD4.F")
markers_long <- imap_dfr(resting_analysis, function(df, nm) {
  df %>%
    as.data.frame() %>%
    mutate(
      cluster = str_remove(nm, "_vs_AllResting$"),
      gene = SYMBOL
    ) %>%
    select(gene, cluster, logFC, AveExpr, P.Value, adj.P.Val, t, B)
}) %>%
  filter(cluster %in% cluster_order)

top10 <- markers_long %>%
  filter(logFC > 0, adj.P.Val < 1e-20) %>%
  group_by(cluster) %>%
  slice_max(order_by = logFC, n = 10, with_ties = FALSE) %>%
  ungroup()

gene_order <- top10 %>%
  mutate(cluster = factor(cluster, levels = cluster_order)) %>%
  arrange(cluster, desc(logFC)) %>%
  distinct(gene, .keep_all = TRUE) %>%
  pull(gene)

lfc_mat_df <- markers_long %>%
  filter(gene %in% gene_order) %>%
  select(gene, cluster, logFC) %>%
  distinct()

lfc_mat <- lfc_mat_df %>%
  pivot_wider(
    names_from = cluster,
    values_from = logFC,
    values_fill = 0
  ) %>%
  as.data.frame()

rownames(lfc_mat) <- lfc_mat$gene
lfc_mat$gene <- NULL

missing_cols <- setdiff(cluster_order, colnames(lfc_mat))
for (cl in missing_cols) {
  lfc_mat[[cl]] <- 0
}

lfc_mat <- lfc_mat[, cluster_order, drop = FALSE]
lfc_mat <- lfc_mat[gene_order, , drop = FALSE]
lfc_mat <- as.matrix(lfc_mat)

lfc_mat[lfc_mat > 2] <- 2
lfc_mat[lfc_mat < -2] <- -2

blue_white_red <- colorRampPalette(c("blue", "white", "red"))(100)

p <- pheatmap(
  lfc_mat,
  color = blue_white_red,
  breaks = seq(-2, 2, length.out = 101),
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  cellwidth = 12,
  cellheight = 12,
  fontsize_row = 6,
  main = "Resting CD4 DE genes (top10 genes, all cluster logFC)"
)

print(p)
```
### Fig 7e
```{r 7e bar plot of percent of cells with expanded clonotypes in resting clusters}
cd4_tcrs <- tcr %>%
  filter(
    level1 == "CD4"
  )


cd4_tcrs <- cd4_tcrs %>%
  filter(
    !is.na(alpha.vgene),
    !is.na(alpha.jgene),
    !is.na(alpha.junction.nt),
    !is.na(beta.vgene),
    !is.na(beta.jgene),
    !is.na(beta.junction.nt)
  ) %>%
  unite(
    "clonotype",
    alpha.vgene,
    alpha.jgene,
    alpha.junction.nt,
    beta.vgene,
    beta.jgene,
    beta.junction.nt,
    sep = ".",
    remove = FALSE
  )



cd4_tcrs$annotation_level2 <-
  cd4_object$annotation_level2[
    match(
      cd4_tcrs$IGT.cellID,
      colnames(cd4_object)
    )
  ]


cd4_tcrs$sample_code <-
  cd4_object$sample_code[
    match(
      cd4_tcrs$IGT.cellID,
      colnames(cd4_object)
    )
  ]


cd4_tcrs$condition_detailed_simplified <-
  cd4_object$condition_detailed_simplified[
    match(
      cd4_tcrs$IGT.cellID,
      colnames(cd4_object)
    )
  ]


cd4_tcrs$organ_simplified <-
  cd4_object$organ_simplified[
    match(
      cd4_tcrs$IGT.cellID,
      colnames(cd4_object)
    )
  ]


cd4_tcrs$target_cells_simplified <-
  cd4_object$target_cells_simplified[
    match(
      cd4_tcrs$IGT.cellID,
      colnames(cd4_object)
    )
  ]


cat(
  "TCR cells matching cd4_object:",
  sum(cd4_tcrs$IGT.cellID %in% colnames(cd4_object)),
  "\n"
)

cat(
  "TCR cells with annotation:",
  sum(!is.na(cd4_tcrs$annotation_level2)),
  "\n"
)


cd4_tcrs_filt <- cd4_tcrs %>%
  filter(
    isMAIT == FALSE,
    isNKT == FALSE
  )


resting_clusters <- c(
  "CD4.A",
  "CD4.B",
  "CD4.C",
  "CD4.D",
  "CD4.E",
  "CD4.F"
)

tcr_plot <- cd4_tcrs_filt %>%
  filter(
    annotation_level2 %in% resting_clusters,
    organ_simplified %in% c(
      "spleen",
      "LN"
    ),
    target_cells_simplified %in% c(
      "allT",
      "CD45p"
    ),
    !is.na(sample_code),
    !is.na(condition_detailed_simplified)
  ) %>%
  mutate(
    condition_group = if_else(
      condition_detailed_simplified == "baseline",
      "Baseline",
      "Challenge"
    )
  )



keep_samples <- tcr_plot %>%
  count(
    sample_code,
    name = "n_cells"
  ) %>%
  filter(
    n_cells > 50
  ) %>%
  pull(sample_code)


tcr_plot <- tcr_plot %>%
  filter(
    sample_code %in% keep_samples
  )


cat(
  "Cells after >50 sample filtering:",
  nrow(tcr_plot),
  "\n"
)

cat(
  "Samples retained:",
  length(unique(tcr_plot$sample_code)),
  "\n"
)

expanded_clonotypes_n2 <- tcr_plot %>%
  count(
    clonotype,
    name = "clonotype_count"
  ) %>%
  filter(
    clonotype_count > 1
  )


cat(
  "Expanded clonotypes:",
  nrow(expanded_clonotypes_n2),
  "\n"
)


tcr_plot <- tcr_plot %>%
  mutate(
    expanded = clonotype %in%
      expanded_clonotypes_n2$clonotype
  )



plot_df <- tcr_plot %>%
  group_by(
    annotation_level2,
    condition_group
  ) %>%
  summarise(
    n_total_cells = n(),
    n_expanded_cells = sum(expanded),
    pct_expanded_cells =
      100 * n_expanded_cells / n_total_cells,
    .groups = "drop"
  )


plot_df$annotation_level2 <- factor(
  plot_df$annotation_level2,
  levels = resting_clusters
)


plot_df$condition_group <- factor(
  plot_df$condition_group,
  levels = c(
    "Baseline",
    "Challenge"
  )
)


#linear scale plot
p <- ggplot(
  plot_df,
  aes(
    x = annotation_level2,
    y = pct_expanded_cells,
    fill = condition_group
  )
) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7,
    color = "black",
    linewidth = 0.2
  ) +
  geom_text(
    aes(
      label = paste0(
        round(
          pct_expanded_cells,
          2
        ),
        "%"
      )
    ),
    position = position_dodge(width = 0.8),
    vjust = -0.4,
    size = 2.5
  ) +
  scale_fill_manual(
    values = c(
      "Baseline" = "black",
      "Challenge" = "red"
    )
  ) +
  scale_y_continuous(
    limits = c(
      0,
      NA
    ),
    expand = expansion(
      mult = c(
        0,
        0.15
      )
    )
  ) +
  labs(
    x = "Resting CD4 cluster",
    y = "% cells with expanded clonotype",
    fill = NULL
  ) +
  theme_bw(
    base_size = 8
  ) +
  theme(
    panel.grid = element_blank(),
    legend.position = "top"
  )


print(p)

```
### Fig 7f
```{r 7f heatmap of conditions in resting clusters}
resting_clusters <- c(
  "CD4.A", "CD4.B", "CD4.C",
  "CD4.D", "CD4.E", "CD4.F"
)

cd4_meta <- cd4_object@meta.data %>%
  filter(
    organ_simplified %in% c("spleen", "LN"),
    target_cells_simplified %in% c("allT", "CD45p"),
    !is.na(sample_code),
    !is.na(condition_detailed_simplified),
    !is.na(annotation_level2)
  )

keep_samples <- cd4_meta %>%
  count(
    sample_code,
    name = "n_cells"
  ) %>%
  filter(
    n_cells >= 50
  ) %>%
  pull(
    sample_code
  )

cd4_meta <- cd4_meta %>%
  filter(
    sample_code %in% keep_samples
  )


condition_totals <- cd4_meta %>%
  count(
    condition_detailed_simplified,
    name = "n_total"
  )


resting_counts <- cd4_meta %>%
  filter(
    annotation_level2 %in% resting_clusters
  ) %>%
  count(
    condition_detailed_simplified,
    annotation_level2,
    name = "n_cluster"
  ) %>%
  complete(
    condition_detailed_simplified,
    annotation_level2 = resting_clusters,
    fill = list(
      n_cluster = 0
    )
  ) %>%
  left_join(
    condition_totals,
    by = "condition_detailed_simplified"
  ) %>%
  mutate(
    proportion = n_cluster / n_total
  )


baseline_name <- "baseline"


baseline_props <- resting_counts %>%
  filter(
    condition_detailed_simplified == baseline_name
  ) %>%
  select(
    annotation_level2,
    baseline_proportion = proportion
  )


pseudocount <- 1e-4


log2fc_counts <- resting_counts %>%
  left_join(
    baseline_props,
    by = "annotation_level2"
  ) %>%
  mutate(
    log2FC = log2(
      (proportion + pseudocount) /
        (baseline_proportion + pseudocount)
    )
  )


heatmap_df <- log2fc_counts %>%
  filter(
    condition_detailed_simplified != baseline_name
  ) %>%
  select(
    condition_detailed_simplified,
    annotation_level2,
    log2FC
  ) %>%
  pivot_wider(
    names_from = annotation_level2,
    values_from = log2FC,
    values_fill = 0
  ) %>%
  column_to_rownames(
    "condition_detailed_simplified"
  ) %>%
  as.data.frame()


heatmap_df <- heatmap_df[
  ,
  resting_clusters,
  drop = FALSE
]


max_abs <- max(
  abs(heatmap_df),
  na.rm = TRUE
)


breaks_use <- seq(
  -max_abs,
  max_abs,
  length.out = 101
)


pheatmap(
  heatmap_df,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  scale = "none",
  border_color = "darkgray",
  color = colorRampPalette(
    c("blue", "white", "red")
  )(100),
  breaks = breaks_use,
  display_numbers = FALSE,
  cellwidth = 12,
  cellheight = 12,
  fontsize_number = 7,
  fontsize_row = 8,
  fontsize_col = 10,
  angle_col = 45,
  main = "log2FC in Proportion vs Baseline"
)

```
# Figure 8

## immgenT CD4 panel: a flow cytometry strategy to resolve the CD4+ T cell landscape

### Fig 8a
```{r 8a dotplot of selected markersfor flow, fig.width=4.375, fig.height=2.5}
gene_markers <- c(
  "Sell", "Cd44", "Cd69" ,"Cxcr5", "Cxcr6", "Izumo1r",
  "Nt5e", "Pdcd1", "Klrg1", "Ly6c1", "Ly6c2", "Il1rl1", "Il23r")

DefaultAssay(cd4_object) <- "RNA"

genes_use <- intersect(gene_markers, rownames(cd4_object[["RNA"]]))
stopifnot(length(genes_use) > 0)

clusters_drop <- c("CD4.wN", "CD4.wZ", "CD4.P", "CD4.wM")

cd4_rna <- subset(
  cd4_object,
  subset = !(annotation_level2 %in% clusters_drop)
)
DefaultAssay(cd4_rna) <- "RNA"

genes_use <- intersect(gene_markers, rownames(cd4_rna[["RNA"]]))
stopifnot(length(genes_use) > 0)

cd4_rna <- NormalizeData(cd4_rna, assay = "RNA", normalization.method = "LogNormalize")
dp <- DotPlot(
  cd4_rna,
  features = genes_use,
  group.by = "annotation_level2",
  assay = "RNA"
)

dp_df <- dp$data
dp_df$features.plot <- factor(dp_df$features.plot, levels = rev(genes_use))

base_size <- 10
axis_size <- 9

p_rna <- ggplot(dp_df, aes(x = id, y = features.plot)) +
  geom_point(
    aes(size = pct.exp, fill = avg.exp.scaled),
    shape = 21,
    color = "black",
    stroke = 0.12
  ) +
  scale_fill_gradient2(
    low = "darkblue",
    mid = "white",
    high = "red",
    midpoint = 0,
    oob = squish,
    name = "Avg expr (scaled)"
  ) +
  scale_size_continuous(
    name = "% expressing",
    range = c(0.5, 4),
    breaks = c(0, 25, 50, 75, 100),
    limits = c(0, 100)
  ) +
  scale_y_discrete(expand = expansion(mult = c(0.04, 0.04))) +
  scale_x_discrete(expand = expansion(mult = c(0.04, 0.04))) +
  theme_bw(base_size = base_size) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(
      size = axis_size,
      angle = 90,
      vjust = 0.5,
      hjust = 1,
      margin = margin(t = 1)
    ),
    axis.text.y = element_text(
      size = axis_size,
      lineheight = 0.85,
      margin = margin(r = 1)
    ),
    panel.grid = element_blank(),
    panel.border = element_rect(linewidth = 0.3),
    plot.margin = margin(2, 2, 2, 2),
    legend.title = element_text(size = base_size),
    legend.text  = element_text(size = base_size - 1)
  ) +
  guides(
    size = guide_legend(title = "% expressing"),
    fill = guide_colorbar(title = "Avg expr (scaled)")
  )
print(p_rna)

```
### Fig 8b
```{r 8b citeseq protein expression dotplot}
marker_order <- c(
  "CD62L", "CD44", "CD69" ,"CXCR5", "CXCR6.CD186", "FR4", "CD73.5NTD",
  "PDCD1.PD1.CD279", "KLRG1", "LY6C1-LY6C2"
)
grp <- factor(cd4_cite$annotation_level2)

adt_data <- GetAssayData(cd4_cite, assay = "ADT", layer = "data")
present_markers <- intersect(marker_order, rownames(adt_data))

adt_avg <- sapply(levels(grp), function(g) {
  cells <- which(grp == g)
  rowMeans(adt_data[present_markers, cells, drop = FALSE])
}) |> as.matrix()

df_plot <- as.data.frame(adt_avg) |>
  rownames_to_column("Marker") |>
  pivot_longer(-Marker, names_to = "Cluster", values_to = "expr") |>
  mutate(across(c(Marker, Cluster), trimws),
         expr = as.numeric(expr)) |>
  group_by(Marker) |>
  mutate(expr_scaled = pmax(pmin(as.numeric(scale(expr)), 2), -2)) |>
  ungroup() |>
  mutate(
    expr_size = pmax(pmin(expr, quantile(expr, 0.99, na.rm = TRUE)) * (1 + expr_scaled / 2), 0.01),
    Marker    = factor(Marker,  levels = rev(present_markers)),
    Cluster   = factor(Cluster, levels = levels(grp))
  )

p_dot <- ggplot(df_plot, aes(x = Cluster, y = Marker)) +
  geom_point(aes(size = expr_size, fill = expr_scaled),
             shape = 21, color = "black", stroke = 0.25) +
  scale_size_continuous(range = c(0.3, 4.5), name = "Expression (raw ADT, z-weighted)") +
  scale_fill_gradient2(low = "darkblue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-2, 2), name = "Expression (z-score; capped)") +
  scale_x_discrete(expand = expansion(mult = c(0.02, 0.02))) +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(title = "ADT markers (scaled color + raw expression size)", x = NULL, y = NULL)

print(p_dot)
```
### Fig 8d
```{r 8d - citeseq gating to highlight clusters, fig.width = 8, fig.height=8}
adt <- GetAssayData(cd4_cite, assay = "ADT", layer = "data")

cd4_cite <- NormalizeData(cd4_cite, assay = "RNA", normalization.method = "LogNormalize")
rna <- GetAssayData(cd4_cite, assay = "RNA", layer = "data")
top10 <- function(x) unname(quantile(x, 0.90, na.rm = TRUE))
top5 <- function(x) unname(quantile(x, 0.95, na.rm = TRUE))
top1 <- function(x) unname(quantile(x, 0.99, na.rm = TRUE))
top2 <- function(x) unname(quantile(x, 0.98, na.rm = TRUE))
top20 <- function(x) unname(quantile(x, 0.80, na.rm = TRUE))
top30 <- function(x) unname(quantile(x, 0.70, na.rm = TRUE))
top50 <- function(x) unname(quantile(x, 0.50, na.rm = TRUE))

cd44_cut  <- top10(adt["CD44", ])
cd44_thresh <- 5
cd62l_cut <- top10(adt["CD62L", ])
cd62l_thresh <- 5
cxcr6_cut <- top1(adt["CXCR6.CD186", ])
cxcr5_cut <- top10(adt["CXCR5", ])
pd1_cut   <- top10(adt["PDCD1.PD1.CD279", ])
ly6c_cut <- top5(adt["LY6C1-LY6C2", ])
st2_cut <- top30(adt["IL1RL1.ST2", ])
klrg1_cut <- top10(adt["KLRG1", ])
klrg1_thresh <- 5
ly6c_thresh <- 4

il23r_vals <- rna["Il23r", , drop = TRUE]
il23r_cut <- top10(il23r_vals)

cxcr6_vals <- rna["Cxcr6", , drop = TRUE]
cxcr5_vals <- rna["Cxcr5", , drop = TRUE]
il1rl1_vals <- rna["Il1rl1", drop = TRUE]


cd4_cite$is_g <-
  (adt["CD44", ]            > cd44_thresh)  &
  (adt["CD62L", ] < cd62l_thresh) &
  (adt["FR4", ] > 4) &
  (adt["CD73.5NTD", ] > 5) &
  (rna["Cxcr6", ]     == 0) &
  (adt["CXCR5", ]           < cxcr5_cut) & #less than top50
  (adt["PDCD1.PD1.CD279", ] < pd1_cut) #less than top50


p <- MyDimPlotHighlightDensity(
  seurat_object = cd4_cite,
  umap_to_plot = "mde_incremental",
  group.by = "is_g",
  split.by = NULL,
 raster = F,
  highlight_size = 0.5,
  highlight_alpha = 1.0 ,
  cols = magma(100)
) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + theme_void()+ NoGrid() + NoLegend()
p 

print(p)



cd4_cite$is_I <-
  (adt["CD44", ]            > cd44_thresh)  &
  (adt["CD62L", ] < cd62l_thresh) &
  (rna["Cxcr6", ]     == 0) &
  (adt["CXCR5", ]         < cxcr5_cut) & #used top1
  (adt["PDCD1.PD1.CD279", ] > pd1_cut) #used top1


p <- MyDimPlotHighlightDensity(
  seurat_object = cd4_cite,
  umap_to_plot = "mde_incremental",
  group.by = "is_I",
  split.by = NULL,
  highlight_size = 0.5,
 raster = F,
  highlight_alpha = 1.0,
  cols = magma(100)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + theme_void() + NoGrid() + NoLegend()


print(p)


cd4_cite$is_H <-
  (adt["CD44", ]            > cd44_thresh)  &
  (adt["CXCR5", ]         > cxcr5_cut) & #top10
  (adt["PDCD1.PD1.CD279", ] > pd1_cut) #top10


p <- MyDimPlotHighlightDensity(
  seurat_object = cd4_cite,
  umap_to_plot = "mde_incremental",
  group.by = "is_H",
  split.by = NULL,
  raster = F,
  highlight_size = 0.5,
  highlight_alpha = 1.0,
  cols = magma(100)
) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + theme_void() +  NoGrid() + NoLegend()


print(p)


 cd4_cite$is_TU <-
  (adt["CD44", ]  > cd44_thresh) &
  (adt["CD62L", ] < cd62l_thresh) &
  (rna["Cxcr6", ] > 0) &
 (adt["KLRG1", ] > klrg1_thresh) &
  (rna["Il1rl1", ] > 0)


p <- MyDimPlotHighlightDensity(
  seurat_object = cd4_cite,
  umap_to_plot = "mde_incremental",
  group.by = "is_TU",
  split.by = NULL,
 raster = F,
  highlight_size = 0.5,
  highlight_alpha = 1.0,
  cols = magma(100)
) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + theme_void()+ NoGrid() + NoLegend()


print(p)

cd4_cite$is_XY <-
  (adt["CD44", ]  > cd44_thresh) &
  (adt["CD62L", ] < cd62l_thresh) &
  (rna["Cxcr6", ] > 0) &
 (adt["IL1RL1.ST2", ] < st2_cut) &
  (adt["LY6C1-LY6C2", ] < ly6c_thresh) &
  (rna["Il23r", ] > 1)

 p<- MyDimPlotHighlightDensity(
  seurat_object = cd4_cite,
  umap_to_plot = "mde_incremental",
  group.by = "is_XY",
  split.by = NULL,
 raster = F,
  highlight_size = 0.5,
  highlight_alpha = 1.0,
  cols = magma(100)
)  + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + theme_void() + NoGrid() + NoLegend()
print(p)


cd4_cite$is_Q <-
  (adt["CD44", ]  > cd44_thresh) &
  (rna["Cxcr6", ] > 0) &
  (adt["CD62L", ] < cd62l_thresh) &
  (rna["Il1rl1", ] == 0) &
  (adt["KLRG1", ] < klrg1_thresh) &
  (adt["LY6C1-LY6C2", ] < ly6c_thresh) &
  (rna["Il23r", ] == 0)

p <- MyDimPlotHighlightDensity(
  seurat_object = cd4_cite,
  umap_to_plot = "mde_incremental",
  group.by = "is_Q",
  split.by = NULL,
  raster = F,
  highlight_size = 0.5,
  highlight_alpha = 1.0,
  cols = magma(100)
) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + theme_void() + NoGrid() + NoLegend()

print(p)

cd4_cite$is_RS <-
  (adt["CD44", ]  > cd44_thresh) &
  (adt["CD62L", ] < cd62l_thresh) &
  (rna["Cxcr6", ] > 0) &
  (rna["Il1rl1", ] == 0) &
   (adt["KLRG1", ] > klrg1_thresh) &
  (adt["LY6C1-LY6C2", ] > ly6c_thresh)

 p <- MyDimPlotHighlightDensity(
  seurat_object = cd4_cite,
  umap_to_plot = "mde_incremental",
  group.by = "is_RS",
  split.by = NULL,
  raster = F,
  highlight_size = 0.5,
  highlight_alpha = 1.0,
  cols = magma(100)
) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + theme_void() + NoGrid() + NoLegend()


print(p)
```
# Extended Data Figure 1

## Chemokine receptor expression across the CD4 landscape

### Fig S1a
```{r S1a - cd44 feature plot in resting cells, fig.width=8, fig.height=8}
cells_cd4_af <- colnames(cd4_object)[
  cd4_object$annotation_level2 %in% clusters_keep
]

MyFeaturePlotHighlight <- function (seurat_object, feature, cells_to_highlight, umap_to_plot = "umap", 
    assay = "RNA", colors = viridis::viridis(10), title = "", 
    pixels = c(512, 512), highlight_size = 1, highlight_alpha = 1, 
    xlim = NULL, ylim = NULL, background_alpha = 0.1) 
{
    dim_data <- seurat_object[[umap_to_plot]]@cell.embeddings
    dim1 <- dim_data[, 1]
    dim2 <- dim_data[, 2]
    tmp_all <- data.frame(seurat_object@meta.data, dim1 = dim1, 
        dim2 = dim2)
    p_base <- ggplot2::ggplot(tmp_all) + scattermore::geom_scattermore(ggplot2::aes(x = dim1, 
        y = dim2), color = "grey", pointsize = 1, alpha = background_alpha, 
        pixels = pixels)
    feature_expr <- Seurat::GetAssayData(seurat_object, assay = assay, 
        layer = "data")[feature, ]
    tmp_highlight <- tmp_all[cells_to_highlight, ]
    tmp_highlight$feature_value <- feature_expr[cells_to_highlight]
    p_final <- p_base + ggplot2::geom_point(data = tmp_highlight, 
        ggplot2::aes(x = dim1, y = dim2, color = feature_value), 
        size = highlight_size, alpha = highlight_alpha) + ggplot2::scale_color_gradientn(colours = colors, 
        name = feature) + ggplot2::theme_void() + NoGrid() + 
        ggplot2::labs(title = title, x = paste0(umap_to_plot, 
            "_1"), y = paste0(umap_to_plot, "_2"))
    if (!is.null(xlim) || !is.null(ylim)) {
        p_final <- p_final + ggplot2::coord_cartesian(xlim = xlim, 
            ylim = ylim)
    }
    return(p_final)
}



DefaultAssay(cd4_object) <- "RNA"
MyFeaturePlotHighlight(
  seurat_object       = cd4_object,
  feature             = "CD44",
  cells_to_highlight  = cells_cd4_af,
  umap_to_plot        = "mde_incremental",
  assay               = "ADT",
  colors              = c("lightgrey", "blue"),
  highlight_size      = 0.5,
  highlight_alpha     = 1,
  background_alpha    = 0.25,
  title               = "CD44 expression in CD4.A-F"
)
```
### Fig S1b
```{r s1b chemokine/cytokine receptor feature plots, fig.width=8, fig.height=8}

FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Cx3cr1"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none") 

FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Ccr7"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none")

FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Cxcr3"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none")

FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Ccr5"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none")

FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Ccr4"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none")
FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Ccr6"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none")

FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Cxcr5"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none")

FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Ccr2"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none")

FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Ccr9"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none")
FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Ccr10"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none") 

FeaturePlot(cd4_object, reduction = "mde_incremental", features = c("Cxcr6"), order = T,  min.cutoff = "q20", max.cutoff = "q99",  cols = c("lightgrey", "blue") , raster = F, raster.dpi = c(512,512)) + coord_cartesian(xlim = c(-2.5, 4), ylim = c(-2.5, 2.5)) + NoAxes() + NoLegend() + theme_void() + theme(legend.position = "none")
```
# Extended Data Figure 2

## Cluster distribution of antigen-specific and clonally expanded cells across immune conditions

### Fig S2a
```{r S2a dotplot of cluster proportions across conditins in cells w/ expanded clonotypes, fig.width= 5.0, fig.height= 4.5}
cd4 <- subset(tcr, subset = level1 == "CD4")
write_csv(cd4, "cd4_tcrs.csv")

cd4_tcrs <- read_csv("cd4_tcrs.csv")
cd4_tcrs <- cd4_tcrs %>%
  filter(
    !is.na(alpha.vgene),
    !is.na(alpha.jgene),
    !is.na(alpha.junction.nt),
    !is.na(beta.vgene),
    !is.na(beta.jgene),
    !is.na(beta.junction.nt)
  ) %>%
  tidyr::unite(
    "clonotype",
    alpha.vgene,
    alpha.jgene,
    alpha.junction.nt,
    beta.vgene,
    beta.jgene,
    beta.junction.nt,
    sep = ".",
    remove = FALSE
  )

expanded_clonotypes <- cd4_tcrs %>%
  count(clonotype, name = "clonotype_count") %>%
  filter(clonotype_count > 9)


expanded_cells <- cd4_tcrs %>%
  semi_join(
    expanded_clonotypes,
    by = "clonotype"
  ) %>%
  pull(IGT.cellID) %>%
  unique()


expanded_cells <- intersect(
  expanded_cells,
  colnames(cd4_object)
)

length(expanded_cells)

cd4_object_expanded <- subset(
  cd4_object,
  cells = expanded_cells)


cd4_object_expanded@meta.data <- cd4_object_expanded@meta.data %>%
  mutate(
    condition_detailed_simplified = trimws(as.character(condition_detailed_simplified)),
    condition_detailed            = trimws(as.character(condition_detailed)),
    condition_broad               = trimws(as.character(condition_broad))
  ) %>%
  mutate(
    condition_detailed_simplified = case_when(
      
      condition_detailed %in% c(
        "12mo", "18mo", "27mo"
      ) ~ "aging",
      
      condition_detailed_simplified %in% c(
        "SFB", "SFB_pregnancy"
      ) ~ "SFB",
      
      condition_detailed_simplified %in% c(
        "PDAC",
        "PDAC_aPD1aCTLA4",
        "PDAC_FishHFD",
        "PDAC_SoyHFD"
      ) ~ "PDAC",
      
      condition_detailed_simplified %in% c(
        "GBMclass",
        "GBMclass_aPD1aCTLA4",
        "GBMmes",
        "GBMmes_aPD1aCTLA4",
        "GBMpron",
        "GBMpron_aPD1aCTLA4"
      ) ~ "GBM",
      
      condition_detailed_simplified %in% c(
        "B16_ACT",
        "B16_ACTaPD1a41BB"
      ) ~ "B16_treated",
      
      condition_detailed_simplified %in% c(
        "HeartTransplant_rejection",
        "HeartTransplant_tolerance"
      ) ~ "HeartTransplant",
      
      TRUE ~ condition_detailed_simplified
    )
  )




cd4_object_expanded <- subset(
  cd4_object_expanded,
  subset = !condition_detailed_simplified %in% c(
    "MHCIIhet_Ea16",
    "MHCIIhet",
    "MHCIIKO_Ea16",
    "MHCIIKO",
    "KbDbKO",
    "KbDbKO_MCMV",
    "KbDbQa1KO",
    "KbDbQa1KO_MCMV",
    "other"
  )
)

# Remove healthy B16 controls
cd4_object_expanded <- subset(
  cd4_object_expanded,
  subset = !(
    condition_detailed_simplified == "B16" &
      condition_broad == "healthy"
  )
)


# Remove pregnancy (confounded by sfb) and EAE
cd4_object_expanded <- subset(
  cd4_object_expanded,
  subset = !(
    condition_detailed_simplified %in% c("pregnancy", "EAE")
  )
)


sample_counts <- table(cd4_object_expanded$sample_name)

keep_samples <- names(
  sample_counts[sample_counts > 10]
)

cd4_filt <- subset(
  cd4_object_expanded,
  subset = sample_name %in% keep_samples
)



cd4_filt@meta.data <- cd4_filt@meta.data %>%
  mutate(
    condition_detailed_simplified =
      trimws(as.character(condition_detailed_simplified))
  )

cond_col    <- "condition_detailed_simplified"
cluster_col <- "annotation_level2"
broad_col   <- "condition_broad"

base_size <- 6.5
axis_size <- 6

cd4_metadata <- cd4_filt@meta.data



broad_order <- c(
  "healthy",
  "bacteria",
  "fungus",
  "virus",
  "parasite",
  "multiinfection",
  "autoimmune",
  "allergy",
  "cancer",
  "immunization",
  "allotransplant"
)

cd4_metadata <- cd4_metadata %>%
  mutate(
    condition_broad = factor(
      condition_broad,
      levels = broad_order
    )
  )



cond_to_broad <- cd4_metadata %>%
  count(
    condition = .data[[cond_col]],
    broad     = .data[[broad_col]],
    name = "cells"
  ) %>%
  group_by(condition) %>%
  slice_max(
    cells,
    n = 1,
    with_ties = FALSE
  ) %>%
  ungroup()



dotplot <- cd4_metadata %>%
  count(
    condition = .data[[cond_col]],
    cluster   = .data[[cluster_col]],
    name = "n"
  ) %>%
  left_join(
    cond_to_broad,
    by = "condition"
  ) %>%
  filter(
    !is.na(condition),
    !is.na(cluster),
    !is.na(broad)
  ) %>%
  group_by(condition) %>%
  mutate(
    total = sum(n),
    prop  = n / total
  ) %>%
  ungroup() %>%
  mutate(
    n_log10 = log10(n + 1),
    broad = factor(
      broad,
      levels = broad_order
    )
  )


cond_levels <- dotplot %>%
  distinct(
    condition,
    broad,
    total
  ) %>%
  arrange(
    broad,
    desc(total),
    condition
  ) %>%
  pull(condition)

dotplot <- dotplot %>%
  mutate(
    condition = factor(
      condition,
      levels = rev(cond_levels)
    )
  )


side <- dotplot %>%
  distinct(
    condition,
    broad
  )

broad_levels <- levels(
  droplevels(dotplot$broad)
)

broad_cols <- c(
  "lightblue4",
  "lightblue1",
  "purple4",
  "green4",
  "yellow3",
  "lightyellow3",
  "pink3",
  "orange",
  "pink4",
  "orange4",
  "darkred"
)

broad_cols <- broad_cols[
  seq_len(length(broad_levels))
]

names(broad_cols) <- broad_levels


p_side <- ggplot(
  side,
  aes(
    x = 1,
    y = condition,
    fill = broad
  )
) +
  geom_tile() +
  scale_fill_manual(
    values = broad_cols,
    drop = FALSE
  ) +
  scale_y_discrete(
    expand = expansion(
      mult = c(0.01, 0.01)
    )
  ) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(
      size = base_size - 1
    ),
    legend.text = element_text(
      size = base_size - 2
    ),
    legend.key.height = unit(
      0.18,
      "cm"
    ),
    legend.key.width = unit(
      0.38,
      "cm"
    ),
    legend.key = element_rect(
      color = "black",
      linewidth = 0.3
    ),
    plot.margin = margin(
      0, 0, 0, 0
    )
  ) +
  guides(
    fill = guide_legend(
      title = "Condition broad",
      nrow = 2,
      byrow = TRUE
    )
  )

legend_broad <- cowplot::get_legend(p_side)

p_side_noleg <- p_side +
  theme(
    legend.position = "none",
    plot.margin = margin(
      2, 0, 2, 0
    )
  )


p_main <- ggplot(
  dotplot,
  aes(
    x = cluster,
    y = condition
  )
) +
  geom_point(
    aes(
      size = n_log10,
      fill = prop
    ),
    shape = 21,
    color = "black",
    stroke = 0.1
  ) +
  scale_fill_gradient(
    low = "white",
    high = "darkgreen",
    limits = c(0, 1),
    oob = scales::squish
  ) +
  scale_size_continuous(
    name = "Expanded cells per dot (log10)",
    range = c(1.5, 4)
  ) +
  theme_bw(
    base_size = base_size
  ) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(
      size = axis_size - 1,
      angle = 90,
      vjust = 0.5,
      hjust = 1
    ),
    axis.text.y = element_text(
      size = axis_size - 1,
      lineheight = 0.75
    ),
    panel.grid = element_blank(),
    panel.border = element_rect(
      linewidth = 0.1
    ),
    plot.margin = margin(
      2, 2, 2, 0
    )
  ) +
  guides(
    size = guide_legend(
      title = "Expanded cells per dot (log10)"
    ),
    fill = guide_colorbar(
      title = "Proportion of expanded cells"
    )
  )




bottom_row <- cowplot::plot_grid(
  p_side_noleg,
  p_main,
  ncol = 2,
  rel_widths = c(0.02, 0.98),
  align = "h",
  axis = "tb"
)

cowplot::plot_grid(
  legend_broad,
  bottom_row,
  ncol = 1,
  rel_heights = c(0.05, 0.95)
)

```
# Extended Data Figure 3

## Polarized clusters co-occur within individual immune responses and exhibit tissue-specific biases

### Fig S3c
```{r S3c dotplot of cluster distribution across organs at baseline, fig.width = 5.0, fig.height=3.5 }
cd4_object@meta.data <- cd4_object@meta.data %>%
  mutate(
    condition_detailed_simplified = trimws(as.character(condition_detailed_simplified)),
    condition_detailed            = trimws(as.character(condition_detailed)),
    condition_broad               = trimws(as.character(condition_broad)),
    organ_simplified              = trimws(as.character(organ_simplified))
  )

cd4_object@meta.data <- cd4_object@meta.data %>%
  mutate(
    condition_detailed_simplified = if_else(
      condition_detailed %in% c("12mo", "18mo", "27mo"),
      "aging",
      condition_detailed_simplified
    )
  )

#  SFB conditions
cd4_object@meta.data <- cd4_object@meta.data %>%
  mutate(
    condition_detailed_simplified = case_when(
      condition_detailed_simplified %in% c("SFB", "SFB_pregnancy") ~ "SFB",
      TRUE ~ condition_detailed_simplified
    )
  )

cd4_object@meta.data <- cd4_object@meta.data %>%
  mutate(
    condition_detailed_simplified = case_when(
      condition_detailed_simplified %in%
        c("PDAC", "PDAC_aPD1aCTLA4", "PDAC_FishHFD", "PDAC_SoyHFD") ~ "PDAC",

      condition_detailed_simplified %in%
        c(
          "GBMclass", "GBMclass_aPD1aCTLA4",
          "GBMmes", "GBMmes_aPD1aCTLA4",
          "GBMpron", "GBMpron_aPD1aCTLA4"
        ) ~ "GBM",

      condition_detailed_simplified %in%
        c("B16_ACT", "B16_ACTaPD1a41BB") ~ "B16 treated" ,

      condition_detailed_simplified %in%
        c(
          "HeartTransplant_rejection",
          "HeartTransplant_tolerance"
        ) ~ "HeartTransplant",

      TRUE ~ condition_detailed_simplified
    )
  )



cd4_object <- subset(
  cd4_object,
  subset = !condition_detailed_simplified %in% c(
    "MHCIIhet_Ea16", "MHCIIhet",
    "MHCIIKO_Ea16", "MHCIIKO",
    "KbDbKO", "KbDbKO_MCMV",
    "KbDbQa1KO", "KbDbQa1KO_MCMV",
    "other"
  )
)

cd4_object <- subset(
  cd4_object,
  subset = !(
    condition_detailed_simplified == "B16" &
      condition_broad == "healthy"
  )
)



sample_counts <- table(cd4_object$sample_name)

keep_samples <- names(
  sample_counts[sample_counts > 10]
)

cd4_filt <- subset(
  cd4_object,
  subset = sample_name %in% keep_samples
)



cd4_healthy <- subset(
  cd4_filt,
  subset = condition_detailed_simplified == "baseline"
)

cd4_healthy@meta.data <- cd4_healthy@meta.data %>%
  mutate(
    organ_simplified = case_when(
      organ_simplified %in% c(
        "small intestine epi",
        "small intestine LP"
      ) ~ "small intestine",

      organ_simplified %in% c(
        "colon epi",
        "colon LP"
      ) ~ "colon",

      TRUE ~ organ_simplified
    )
  )

org_col     <- "organ_simplified"
cluster_col <- "annotation_level2"

cd4_metadata <- cd4_healthy@meta.data


dotplot <- cd4_metadata %>%
  count(
    organ   = .data[[org_col]],
    cluster = .data[[cluster_col]],
    name = "n"
  ) %>%
  filter(
    !is.na(organ),
    !is.na(cluster)
  ) %>%
  group_by(organ) %>%
  mutate(
    total = sum(n),
    prop  = n / total
  ) %>%
  ungroup() %>%
  mutate(
    n_log10 = log10(n + 1)
  )



organ_levels <- dotplot %>%
  distinct(organ, total) %>%
  arrange(desc(total), organ) %>%
  pull(organ)

dotplot <- dotplot %>%
  mutate(
    organ = factor(
      organ,
      levels = rev(organ_levels)
    )
  )


p_main <- ggplot(
  dotplot,
  aes(
    x = cluster,
    y = organ
  )
) +
  geom_point(
    aes(
      size = n_log10,
      fill = prop
    ),
    shape = 21,
    color = "black",
    stroke = 0.1
  ) +
  scale_fill_gradient(
    low = "white",
    high = "darkgreen",
    limits = c(0, 0.25),
    oob = scales::squish
  ) +
  scale_size_continuous(
    name = "Cells per dot (log10)",
    range = c(1.5, 4)
  ) +
  theme_bw(
    base_size = base_size
  ) +
  theme(
    axis.title = element_blank(),

    axis.text.x = element_text(
      size = axis_size - 1,
      angle = 90,
      vjust = 0.5,
      hjust = 1
    ),

    axis.text.y = element_text(
      size = axis_size - 1,
      lineheight = 0.75
    ),

    panel.grid = element_blank(),

    panel.border = element_rect(
      linewidth = 0.1
    ),

    plot.margin = margin(
      2, 2, 2, 2
    )
  ) +
  guides(
    size = guide_legend(
      title = "Cells per dot (log10)"
    ),
    fill = guide_colorbar(
      title = "Proportion"
    )
  )


print(p_main)
```
### Fig S3d
```{r S3d_log10_cd4s_spleen_LN_clusters, fig.width=12, fig.height=5}
organ_meta <- cd4_object@meta.data %>%
  filter(
    condition_detailed_simplified == "baseline",
    organ_simplified %in% c("spleen", "LN"),
    !is.na(annotation_level2),
    !is.na(IGTHT)
  ) %>%
  mutate(
    organ_simplified = factor(
      organ_simplified,
      levels = c("spleen", "LN")
    )
  ) %>%
  group_by(
    IGTHT,
    organ_simplified
  ) %>%
  filter(
    n() >= 50
  ) %>%
  ungroup()


cluster_order <- levels(
  cd4_object$annotation_level2
)

cluster_order <- cluster_order[
  cluster_order %in%
    unique(as.character(organ_meta$annotation_level2))
]


igtht_prop <- organ_meta %>%
  count(
    IGTHT,
    organ_simplified,
    annotation_level2,
    name = "n_cluster"
  ) %>%
  group_by(
    IGTHT,
    organ_simplified
  ) %>%
  complete(
    annotation_level2 = cluster_order,
    fill = list(
      n_cluster = 0
    )
  ) %>%
  mutate(
    n_total = sum(n_cluster),
    proportion = n_cluster / n_total,
    percent = 100 * proportion
  ) %>%
  ungroup() %>%
  mutate(
    annotation_level2 = factor(
      annotation_level2,
      levels = cluster_order
    ),
    organ_simplified = factor(
      organ_simplified,
      levels = c("spleen", "LN")
    )
  )


igtht_prop_log <- igtht_prop %>%
  mutate(
    percent_plot = if_else(
      percent > 0,
      percent,
      NA_real_
    )
  )
p_organ_prop_log <- ggplot(
  igtht_prop_log,
  aes(
    x = organ_simplified,
    y = percent_plot,
    color = annotation_level2
  )
) +
  geom_jitter(
    width = 0.15,
    height = 0,
    size = 1.5,
    alpha = 0.8,
    na.rm = TRUE
  ) +
  stat_summary(
    fun = mean,
    geom = "crossbar",
    width = 0.4,
    color = "black",
    linewidth = 0.7,
    na.rm = TRUE
  ) +
  scale_color_manual(
    values = palette_vector
  ) +
scale_y_log10(
  breaks = c(
    0.1,
    1,
    10,
    100
  ),
  minor_breaks = scales::minor_breaks_log(),
  labels = c(
    "0.1%",
    "1%",
    "10%",
    "100%"
  ),
  expand = expansion(
    mult = c(
      0.02,
      0.05
    )
  )
) +
  guides(
    y = guide_axis(
      minor.ticks = TRUE
    )
  ) +
  labs(
    x = NULL,
    y = "% among all CD4 T cells (log10 scale)"
  ) +
  facet_grid(
    . ~ annotation_level2,
    scales = "free_x",
    space = "free_x",
    switch = "x",
    drop = TRUE
  ) +
  theme_classic() +
  theme(
    legend.position = "none",

    strip.background = element_blank(),
    strip.placement = "outside",

    strip.text.x = element_text(
      size = 10,
      margin = margin(t = 3)
    ),

    panel.spacing.x = unit(
      0.15,
      "lines"
    ),

    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    ),

    axis.title.x = element_blank(),

    axis.ticks.y = element_line(
      linewidth = 0.5
    ),

    axis.minor.ticks.y.left = element_line(
      linewidth = 0.35
    ),

    axis.ticks.length.y = unit(
      5,
      "pt"
    ),

    axis.minor.ticks.length.y = unit(
      2.5,
      "pt"
    )
  )


p_organ_prop_log
```
### Fig S3e
```{r S3e barplot of cluster distribution in ifng+ spleen/ln cells, fig.width=10, fig.height=5}

DefaultAssay(cd4_object) <- "RNA"

level2_order <- levels(cd4_object$annotation_level2)

level2_order <- level2_order[
  grepl("^CD4\\.", level2_order)
]

ifng_expr <- FetchData(
  cd4_object,
  vars = "Ifng",
  layer = "data"
)

ifng_cells <- colnames(cd4_object)[
  cd4_object$organ_simplified %in% c("spleen", "LN") &
    ifng_expr$Ifng > 0 &
    !is.na(cd4_object$annotation_level2)
]

ifng_counts <- cd4_object@meta.data[
  ifng_cells,
  ,
  drop = FALSE
] %>%
  filter(
    annotation_level2 %in% level2_order
  ) %>%
  count(
    organ_simplified,
    annotation_level2,
    name = "n_cells"
  ) %>%
  mutate(
    annotation_level2 = as.character(annotation_level2),
    organ_simplified = as.character(organ_simplified)
  )

ifng_organ_df <- expand_grid(
  annotation_level2 = level2_order,
  organ_simplified = c("spleen", "LN")
) %>%
  left_join(
    ifng_counts,
    by = c(
      "annotation_level2",
      "organ_simplified"
    )
  ) %>%
  mutate(
    n_cells = replace_na(n_cells, 0)
  ) %>%
  group_by(organ_simplified) %>%
  mutate(
    percent = 100 * n_cells / sum(n_cells)
  ) %>%
  ungroup() %>%
  mutate(
    annotation_level2 = factor(
      annotation_level2,
      levels = level2_order
    ),
    organ_simplified = factor(
      organ_simplified,
      levels = c("spleen", "LN")
    )
  )

p_ifng_organ <- ggplot(
  ifng_organ_df,
  aes(
    x = annotation_level2,
    y = percent,
    fill = annotation_level2,
    group = organ_simplified
  )
) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7,
    color = "black",
    linewidth = 0.25
  ) +
  geom_text(
    aes(
      y = -0.5,
      label = organ_simplified,
      group = organ_simplified
    ),
    position = position_dodge(width = 0.8),
    angle = 90,
    hjust = 1,
    size = 2.5
  ) +
  scale_fill_manual(
    values = palette_vector,
    drop = FALSE
  ) +
  scale_x_discrete(
    drop = FALSE
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0.12, 0.08))
  ) +
  coord_cartesian(
    clip = "off"
  ) +
  labs(
    x = NULL,
    y = "% of Ifng+ cells"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    legend.position = "none",
    plot.margin = margin(
      t = 5,
      r = 5,
      b = 25,
      l = 5
    )
  )

p_ifng_organ
```
# Extended Data Figure 4
## In vivo Th signature expression is conserved across tissues and varies in intensity and combination across clusters at the single-cell level
### Fig S4a-d

```{r s4a-d polarized cluster signatureheatmaps across organs, fig.width=6, fig.height=15}
gene_lists <- list(
  XY = xy_genes,
  RS = rs_genes,
  TU = tu_genes,
  HI = hi_genes
)
cd4_use <- cd4_filt

cd4_use@meta.data <- cd4_use@meta.data %>%
  mutate(
    organ_simplified = case_when(
      organ_simplified %in% c(
        "small intestine epi",
        "small intestine LP"
      ) ~ "small intestine",

      organ_simplified %in% c(
        "colon epi",
        "colon LP"
      ) ~ "colon",

      TRUE ~ organ_simplified
    ),

    polarized_group = case_when(
      annotation_level2 %in% c(
        "CD4.X",
        "CD4.Y"
      ) ~ "XY",

      annotation_level2 %in% c(
        "CD4.R",
        "CD4.S"
      ) ~ "RS",

      annotation_level2 %in% c(
        "CD4.T",
        "CD4.U"
      ) ~ "TU",

      annotation_level2 %in% c(
        "CD4.H",
        "CD4.I"
      ) ~ "HI",

      TRUE ~ NA_character_
    )
  )


heatmap_results <- list()


for (polarized_name in c("XY", "RS", "TU", "HI")) {

  message("Running ", polarized_name)

  gene_list <- gene_lists[[polarized_name]]


  meta <- cd4_use@meta.data %>%
    mutate(
      cell = rownames(.)
    ) %>%
    filter(
      polarized_group == polarized_name,
      condition_detailed_simplified == "baseline",
      !is.na(organ_simplified),
      !is.na(IGTHT)
    )


  pb_counts <- meta %>%
    count(
      organ_simplified,
      name = "n_cells"
    ) %>%
    filter(
      n_cells >= 20
    )


  meta <- meta %>%
    semi_join(
      pb_counts,
      by = "organ_simplified"
    )


  if (nrow(meta) == 0) {

    message(
      polarized_name,
      ": no pseudobulks with >= 20 cells; skipping"
    )

    next
  }


  pb_meta <- meta %>%
    distinct(
      organ_simplified
    ) %>%
    arrange(
      organ_simplified
    ) %>%
    mutate(
      pseudobulk_id = sprintf(
        "PB%03d",
        row_number()
      )
    ) %>%
    left_join(
      pb_counts,
      by = "organ_simplified"
    )


  meta <- meta %>%
    left_join(
      pb_meta %>%
        select(
          organ_simplified,
          pseudobulk_id
        ),
      by = "organ_simplified"
    )


  obj_use <- subset(
    cd4_use,
    cells = meta$cell
  )


  meta <- meta[
    match(
      colnames(obj_use),
      meta$cell
    ),
    ,
    drop = FALSE
  ]


  obj_use$pseudobulk_id <- meta$pseudobulk_id


  if ("Assay5" %in% class(obj_use[["RNA"]])) {

    obj_use[["RNA"]] <- JoinLayers(
      obj_use[["RNA"]]
    )
  }


  counts_mat <- GetAssayData(
    obj_use,
    assay = "RNA",
    layer = "counts"
  )


  pb_factor <- factor(
    obj_use$pseudobulk_id,
    levels = pb_meta$pseudobulk_id
  )


  design_indicator <- Matrix::sparse.model.matrix(
    ~ 0 + pb_factor
  )


  colnames(design_indicator) <- levels(
    pb_factor
  )


  pb_raw_counts <- counts_mat %*%
    design_indicator


  pb_raw_counts <- as.matrix(
    pb_raw_counts
  )


  pb_meta <- pb_meta[
    match(
      colnames(pb_raw_counts),
      pb_meta$pseudobulk_id
    ),
    ,
    drop = FALSE
  ]


  library_sizes <- colSums(
    pb_raw_counts
  )


  cp10k <- sweep(
    pb_raw_counts,
    2,
    library_sizes,
    "/"
  ) * 10000


  log1p_cp10k <- log1p(
    cp10k
  )


  genes_keep <- gene_list[
    gene_list %in% rownames(log1p_cp10k)
  ]


  if (length(genes_keep) == 0) {

    message(
      polarized_name,
      ": none of the signature genes are present; skipping"
    )

    next
  }


  log1p_cp10k_sig <- log1p_cp10k[
    genes_keep,
    ,
    drop = FALSE
  ]


  organ_levels <- levels(
    factor(cd4_use$organ_simplified)
  )


  organs <- organ_levels[
    organ_levels %in%
      as.character(pb_meta$organ_simplified)
  ]


  if (length(organs) == 0) {

    message(
      polarized_name,
      ": no organs available after filtering; skipping"
    )

    next
  }


  organ_list <- lapply(
    organs,
    function(org) {

      cols_use <- pb_meta$pseudobulk_id[
        pb_meta$organ_simplified == org
      ]


      if (length(cols_use) == 1) {

        vals <- log1p_cp10k_sig[
          genes_keep,
          cols_use,
          drop = TRUE
        ]

      } else {

        vals <- rowMeans(
          log1p_cp10k_sig[
            genes_keep,
            cols_use,
            drop = FALSE
          ]
        )
      }


      vals <- as.numeric(vals)

      names(vals) <- genes_keep

      vals
    }
  )


  organ_mat <- do.call(
    cbind,
    organ_list
  )


  organ_mat <- as.matrix(
    organ_mat
  )


  rownames(organ_mat) <- genes_keep

  colnames(organ_mat) <- organs


  organ_counts <- pb_meta %>%
    group_by(
      organ_simplified
    ) %>%
    summarise(
      total_cells = sum(n_cells),
      .groups = "drop"
    )


  organ_counts <- organ_counts[
    match(
      colnames(organ_mat),
      organ_counts$organ_simplified
    ),
    ,
    drop = FALSE
  ]


  colnames(organ_mat) <- paste0(
    organ_counts$organ_simplified,
    " (n=",
    organ_counts$total_cells,
    ")"
  )


  pheatmap(
    organ_mat,

    color = colorRampPalette(
      c(
        "white",
        "red"
      )
    )(100),

    scale = "none",

    cluster_rows = FALSE,
    cluster_cols = FALSE,

    border_color = "darkgray",

    fontsize = 7,
    fontsize_row = 5,
    fontsize_col = 7,

    cellwidth = 12,
    cellheight = 12,

    angle_col = 90,

    main = paste0(
      polarized_name,
      " signature\nmean log1p(CP10K), baseline"
    )
  )


  heatmap_results[[polarized_name]] <- list(
    pseudobulk_metadata = pb_meta,
    pseudobulk_cell_counts = pb_counts,
    raw_pseudobulk_counts = pb_raw_counts,
    cp10k = cp10k,
    log1p_cp10k = log1p_cp10k,
    heatmap_matrix = organ_mat
  )
}
```
### S4e-h
```{r s4e-h scatter plots of module scores plotting all clusters, fig.width = 8, fig.height=8}
clusters_keep <- c("CD4.Q","CD4.G","CD4.H", "CD4.I", "CD4.R", "CD4.S", "CD4.T", "CD4.U", "CD4.X", "CD4.Y")

modulescores <- FetchData(
  cd4_object,
  vars = c("TH1_Score1", "TH2_Score1", "TH17_Score1", "TFH_TCHRON_Score1", "annotation_level2")
) %>%
  filter(annotation_level2 %in% clusters_keep)

modulescores <- modulescores %>%
  mutate(
    TH1_Score1 = pmin(TH1_Score1, 2.0),
    TH2_Score1 = pmin(TH2_Score1, 2.0),
    TH17_Score1 = pmin(TH17_Score1, 2.0),
    TFH_TCHRON_Score1 = pmin(TFH_TCHRON_Score1, 2.0)
  )

mean <- modulescores %>%
  group_by(annotation_level2) %>%
  summarise(
    TH1 = mean(TH1_Score1, na.rm = TRUE),
    TH2 = mean(TH2_Score1, na.rm = TRUE),
    TH17 = mean(TH17_Score1, na.rm = TRUE),
    TFH_TCHRON = mean(TFH_TCHRON_Score1, na.rm = TRUE)
  )
#th1 vs th2
quadrant_pct <- modulescores %>%
  mutate(quadrant = case_when(
    TH2_Score1 >= 0.25 & TH1_Score1 >= 0 ~ "Q1 (TH2+TH1+)",
    TH2_Score1 <  0.25 & TH1_Score1 >= 0 ~ "Q2 (TH2-TH1+)",
    TH2_Score1 <  0.25 & TH1_Score1 <  0 ~ "Q3 (TH2-TH1-)",
    TH2_Score1 >= 0.25 & TH1_Score1 <  0 ~ "Q4 (TH2+TH1-)"
  )) %>%
  count(quadrant) %>%
  mutate(pct = n / sum(n) * 100,
         label = sprintf("%.1f%%", pct))

quadrant_coords <- data.frame(
  quadrant = c("Q1 (TH2+TH1+)", "Q2 (TH2-TH1+)", "Q3 (TH2-TH1-)", "Q4 (TH2+TH1-)"),
  x = c(1.2,  -0.1,  -0.1,  1.2),
  y = c(1.0,   1.3,  -0.2,  -0.2)
)

quadrant_pct <- left_join(quadrant_pct, quadrant_coords, by = "quadrant")

ggplot(modulescores, aes(x = TH2_Score1, y = TH1_Score1, color = annotation_level2)) +
  geom_point_rast(size = 0.4, alpha = 0.3, raster.dpi = 200) + 
  geom_point(
    data = mean,
    aes(x = TH2, y = TH1, fill = annotation_level2),
    shape = 21, color = "black", size = 4, stroke = 1
  ) + 
  geom_text(
    data = mean,
    aes(x = TH2, y = TH1, label = annotation_level2),
    color = "black", vjust = -1, size = 4
  ) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", linewidth = 0.4) +
  geom_vline(xintercept = 0.25, color = "red", linetype = "dashed", linewidth = 0.4) +
  
  geom_text(
    data = quadrant_pct,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    color = "black", fontface = "bold", size = 4
  ) +
  theme_classic() +
  labs(x = "Th2-like module score", y = "Th1-like module score",
       color = "Cluster", fill = "Cluster") +
  coord_cartesian(xlim = c(-0.2, 1.5), ylim = c(-0.5, 1.5)) +
  scale_color_manual(values = palette_vector) +
  scale_fill_manual(values = palette_vector)



#th1 vs th17
quadrant_pct <- modulescores %>%
  mutate(quadrant = case_when(
    TH17_Score1 >= 0.25 & TH1_Score1 >= 0 ~ "Q1 (TH17+TH1+)",
    TH17_Score1 <  0.25 & TH1_Score1 >= 0 ~ "Q2 (TH17-TH1+)",
    TH17_Score1 <  0.25 & TH1_Score1 <  0 ~ "Q3 (TH17-TH1-)",
    TH17_Score1 >= 0.25 & TH1_Score1 <  0 ~ "Q4 (TH17+TH1-)"
  )) %>%
  count(quadrant) %>%
  mutate(pct = n / sum(n) * 100,
         label = sprintf("%.1f%%", pct))

quadrant_coords <- data.frame(
  quadrant = c("Q1 (TH17+TH1+)", "Q2 (TH17-TH1+)", "Q3 (TH17-TH1-)", "Q4 (TH17+TH1-)"),
  x = c(1.2,  -0.1,  -0.1,  1.2),
  y = c(1.0,   1.3,  -0.2,  -0.2)
)


quadrant_pct <- left_join(quadrant_pct, quadrant_coords, by = "quadrant")

ggplot(modulescores, aes(x = TH17_Score1, y = TH1_Score1, color = annotation_level2)) +
  geom_point_rast(size = 0.4, alpha = 0.3, raster.dpi = 200) + 
  geom_point(
    data = mean,
    aes(x = TH17, y = TH1, fill = annotation_level2),
    shape = 21, color = "black", size = 4, stroke = 1
  ) + 
  geom_text(
    data = mean,
    aes(x = TH17, y = TH1, label = annotation_level2),
    color = "black", vjust = -1, size = 4
  ) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", linewidth = 0.4) +
  geom_vline(xintercept = 0.25, color = "red", linetype = "dashed", linewidth = 0.4) +
  
  geom_text(
    data = quadrant_pct,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    color = "black", fontface = "bold", size = 4
  ) +
  theme_classic() +
  labs(x = "Th17-like module score", y = "Th1-like module score",
       color = "Cluster", fill = "Cluster") +
  coord_cartesian(xlim = c(-0.3, 1.5), ylim = c(-0.5, 1.5)) +
  scale_color_manual(values = palette_vector) +
  scale_fill_manual(values = palette_vector)



#th2 vs th17
quadrant_pct <- modulescores %>%
  mutate(quadrant = case_when(
    TH17_Score1 >= 0.25 & TH2_Score1 >= 0.25 ~ "Q1 (TH17+TH2+)",
    TH17_Score1 <  0.25 & TH2_Score1 >= 0.25 ~ "Q2 (TH17-TH2+)",
    TH17_Score1 <  0.25 & TH2_Score1 <  0.25 ~ "Q3 (TH17-TH2-)",
    TH17_Score1 >= 0.25 & TH2_Score1 <  0.25 ~ "Q4 (TH17+TH2-)"
  )) %>%
  count(quadrant) %>%
  mutate(pct = n / sum(n) * 100,
         label = sprintf("%.1f%%", pct))

quadrant_coords <- data.frame(
  quadrant = c("Q1 (TH17+TH2+)", "Q2 (TH17-TH2+)", "Q3 (TH17-TH2-)", "Q4 (TH17+TH2-)"),
  x = c(1.2,  -0.1,  -0.1,  1.2),
  y = c(1.0,   1.3,  -0.2,  -0.2)
)


quadrant_pct <- left_join(quadrant_pct, quadrant_coords, by = "quadrant")

ggplot(modulescores, aes(x = TH17_Score1, y = TH2_Score1, color = annotation_level2)) +
  geom_point_rast(size = 0.4, alpha = 0.3, raster.dpi = 200) + 
  geom_point(
    data = mean,
    aes(x = TH17, y = TH2, fill = annotation_level2),
    shape = 21, color = "black", size = 4, stroke = 1
  ) + 
  geom_text(
    data = mean,
    aes(x = TH17, y = TH2, label = annotation_level2),
    color = "black", vjust = -1, size = 4
  ) +
  geom_hline(yintercept = 0.25, color = "red", linetype = "dashed", linewidth = 0.4) +
  geom_vline(xintercept = 0.25, color = "red", linetype = "dashed", linewidth = 0.4) +
  
  geom_text(
    data = quadrant_pct,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    color = "black", fontface = "bold", size = 4
  ) +
  theme_classic() +
  labs(x = "Th17-like module score", y = "Th2-like module score",
       color = "Cluster", fill = "Cluster") +
  coord_cartesian(xlim = c(-0.3, 1.5), ylim = c(-0.2, 1.5)) +
  scale_color_manual(values = palette_vector) +
  scale_fill_manual(values = palette_vector)

#tfhtchron vs th1
quadrant_pct <- modulescores %>%
  mutate(quadrant = case_when(
    TFH_TCHRON_Score1 >= 0.25 & TH1_Score1 >= 0 ~ "Q1 (TFH_TCHRON+TH1+)",
    TFH_TCHRON_Score1 <  0.25 & TH1_Score1 >= 0 ~ "Q2 (TFH_TCHRON-TH1+)",
    TFH_TCHRON_Score1 <  0.25 & TH1_Score1 <  0 ~ "Q3 (TFH_TCHRON-TH1-)",
    TFH_TCHRON_Score1 >= 0.25 & TH1_Score1 <  0 ~ "Q4 (TFH_TCHRON+TH1-)"
  )) %>%
  count(quadrant) %>%
  mutate(pct = n / sum(n) * 100,
         label = sprintf("%.1f%%", pct))

quadrant_coords <- data.frame(
  quadrant = c("Q1 (TFH_TCHRON+TH1+)", "Q2 (TFH_TCHRON-TH1+)", "Q3 (TFH_TCHRON-TH1-)", "Q4 (TFH_TCHRON+TH1-)"),
  x = c(1.2,  -0.1,  -0.1,  1.2),
  y = c(1.0,   1.3,  -0.2,  -0.2)
)

quadrant_pct <- left_join(quadrant_pct, quadrant_coords, by = "quadrant")

ggplot(modulescores, aes(x = TFH_TCHRON_Score1, y = TH1_Score1, color = annotation_level2)) +
  geom_point_rast(size = 0.4, alpha = 0.3, raster.dpi = 200) + 
  geom_point(
    data = mean,
    aes(x = TFH_TCHRON, y = TH1, fill = annotation_level2),
    shape = 21, color = "black", size = 4, stroke = 1
  ) + 
  geom_text(
    data = mean,
    aes(x = TFH_TCHRON, y = TH1, label = annotation_level2),
    color = "black", vjust = -1, size = 4
  ) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", linewidth = 0.4) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", linewidth = 0.4) +
  
  geom_text(
    data = quadrant_pct,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    color = "black", fontface = "bold", size = 4
  ) +
  theme_classic() +
  labs(x = "Tfh/Tchron-like module score", y = "Th1-like module score",
       color = "Cluster", fill = "Cluster") +
  coord_cartesian(xlim = c(-0.50, 1.5), ylim = c(-0.5, 1.5)) +
  scale_color_manual(values = palette_vector) +
  scale_fill_manual(values = palette_vector)
```
### S4i-k
```{r S4i-k volcano plots comparing Qth1 vs Qth2 vs Qth17} 

# ORIGINAL DGE GENERATION
# Commented out because the precomputed DGE is loaded below

# xy_genes <- readLines("XY_vs_RSTUHI_up.txt")
# rs_genes <- readLines("RS_vs_TUXYHI_up.txt")
# tu_genes <- readLines("TU_vs_RSXYHI_up.txt")
# hi_genes <- readLines("HI_vs_RSTUXY_up.txt")
#
# DefaultAssay(cd4_object) <- "RNA"
#
# if ("Assay5" %in% class(cd4_object[["RNA"]])) {
#   cd4_object[["RNA"]] <- JoinLayers(cd4_object[["RNA"]])
# }
#
# if (!"TH1_Score1" %in% colnames(cd4_object@meta.data)) {
#   cd4_object <- AddModuleScore(
#     object = cd4_object,
#     features = list(rs_genes),
#     name = "TH1_Score"
#   )
# }
#
# if (!"TH2_Score1" %in% colnames(cd4_object@meta.data)) {
#   cd4_object <- AddModuleScore(
#     object = cd4_object,
#     features = list(tu_genes),
#     name = "TH2_Score"
#   )
# }
#
# if (!"TH17_Score1" %in% colnames(cd4_object@meta.data)) {
#   cd4_object <- AddModuleScore(
#     object = cd4_object,
#     features = list(xy_genes),
#     name = "TH17_Score"
#   )
# }
#
# if (!"TFH_TCHRON_Score1" %in% colnames(cd4_object@meta.data)) {
#   cd4_object <- AddModuleScore(
#     object = cd4_object,
#     features = list(hi_genes),
#     name = "TFH_TCHRON_Score"
#   )
# }
#
# polyfunctional <- readRDS(
#   "CD4_GHIQRSTUXY_igt1_96_withtotalvi20260206_clean_withThPrograms.Rds"
# )
#
# prog_df <- polyfunctional@meta.data %>%
#   select(annotation_level2_program) %>%
#   rownames_to_column("cell") %>%
#   distinct(
#     cell,
#     .keep_all = TRUE
#   ) %>%
#   mutate(
#     annotation_level2_program_clean =
#       tolower(annotation_level2_program)
#   )
#
# q_no_module_thresh <- 0
#
# meta_full <- cd4_object@meta.data %>%
#   rownames_to_column("cell") %>%
#   left_join(
#     prog_df,
#     by = "cell"
#   ) %>%
#   mutate(
#     q_max_program = pmax(
#       TH1_Score1,
#       TH2_Score1,
#       TH17_Score1,
#       TFH_TCHRON_Score1,
#       na.rm = TRUE
#     ),
#
#     is_q_single_or_none =
#       annotation_level2_program_clean %in%
#       c(
#         "cd4.q.none",
#         "cd4.q.th1",
#         "cd4.q.th2",
#         "cd4.q.th17",
#         "cd4.q.tfhchron",
#         "cd4.q.tfh_tchron"
#       ),
#
#     is_q_polyfunctional =
#       annotation_level2 == "CD4.Q" &
#       !is.na(annotation_level2_program_clean) &
#       grepl(
#         "^cd4\\.q\\.",
#         annotation_level2_program_clean
#       ) &
#       !is_q_single_or_none,
#
#     q_program = case_when(
#       annotation_level2 != "CD4.Q" ~
#         NA_character_,
#
#       is_q_polyfunctional ~
#         "CD4.Q_polyfunctional",
#
#       q_max_program < q_no_module_thresh ~
#         "CD4.Q_none",
#
#       TH1_Score1 >= TH2_Score1 &
#         TH1_Score1 >= TH17_Score1 &
#         TH1_Score1 >= TFH_TCHRON_Score1 ~
#         "CD4.Q_th1",
#
#       TH2_Score1 >= TH1_Score1 &
#         TH2_Score1 >= TH17_Score1 &
#         TH2_Score1 >= TFH_TCHRON_Score1 ~
#         "CD4.Q_th2",
#
#       TH17_Score1 >= TH1_Score1 &
#         TH17_Score1 >= TH2_Score1 &
#         TH17_Score1 >= TFH_TCHRON_Score1 ~
#         "CD4.Q_th17",
#
#       TFH_TCHRON_Score1 >= TH1_Score1 &
#         TFH_TCHRON_Score1 >= TH2_Score1 &
#         TFH_TCHRON_Score1 >= TH17_Score1 ~
#         "CD4.Q_tfh_tchron",
#
#       TRUE ~
#         "CD4.Q_none"
#     )
#   )
#
# rownames(meta_full) <- meta_full$cell
# meta_full$cell <- NULL
#
# cd4_object@meta.data <- meta_full
#
# q_compare <- subset(
#   cd4_object,
#   subset = q_program %in%
#     c(
#       "CD4.Q_th1",
#       "CD4.Q_th2",
#       "CD4.Q_th17"
#     )
# )
#
# q_compare$q_program <- factor(
#   q_compare$q_program,
#   levels = c(
#     "CD4.Q_th1",
#     "CD4.Q_th2",
#     "CD4.Q_th17"
#   )
# )
#
# Idents(q_compare) <- "q_program"
#
#
# make_q_dge <- function(
#     object,
#     group1,
#     group2,
#     signatures_to_label
# ) {
#
#   de <- FindMarkers(
#     object = object,
#     ident.1 = group1,
#     ident.2 = group2,
#     assay = "RNA",
#     slot = "data",
#     test.use = "wilcox",
#     logfc.threshold = 0,
#     min.pct = 0,
#     only.pos = FALSE,
#     verbose = FALSE
#   )
#
#   de <- de %>%
#     rownames_to_column("gene") %>%
#     mutate(
#       neglog10_padj = -log10(
#         pmax(
#           p_val_adj,
#           1e-300
#         )
#       ),
#
#       signature = case_when(
#         gene %in% rs_genes ~ "Th1",
#         gene %in% tu_genes ~ "Th2",
#         gene %in% xy_genes ~ "Th17",
#         TRUE ~ "Other"
#       )
#     )
#
#   return(de)
# }
#
#
# q_th1_vs_th2 <- make_q_dge(
#   object = q_compare,
#   group1 = "CD4.Q_th1",
#   group2 = "CD4.Q_th2"
# )
#
# q_th1_vs_th17 <- make_q_dge(
#   object = q_compare,
#   group1 = "CD4.Q_th1",
#   group2 = "CD4.Q_th17"
# )
#
# q_th2_vs_th17 <- make_q_dge(
#   object = q_compare,
#   group1 = "CD4.Q_th2",
#   group2 = "CD4.Q_th17"
# )
#
#
# q_volcano_dge_all <- bind_rows(
#   q_th1_vs_th2 %>%
#     mutate(
#       comparison = "CD4.Q_th1_vs_CD4.Q_th2"
#     ),
#
#   q_th1_vs_th17 %>%
#     mutate(
#       comparison = "CD4.Q_th1_vs_CD4.Q_th17"
#     ),
#
#   q_th2_vs_th17 %>%
#     mutate(
#       comparison = "CD4.Q_th2_vs_CD4.Q_th17"
#     )
# )
#
# write.csv(
#   q_volcano_dge_all,
#   "Q_program_volcano_DGE_all.csv",
#   row.names = FALSE
# )


# LOADING PRECOMPUTED DGE AND GENERATE VOLCANO PLOTS
signature_colors <- c(
  "Th1" = "#7e4e99",
  "Th2" = "#da488c",
  "Th17" = "#eed7b7"
)


comparison_info <- list(
  "CD4.Q_th1_vs_CD4.Q_th2" = list(
    group1 = "CD4.Q_th1",
    group2 = "CD4.Q_th2",
    signatures = c(
      "Th1",
      "Th2"
    ),
    filename = "QTh1_vs_QTh2"
  ),

  "CD4.Q_th1_vs_CD4.Q_th17" = list(
    group1 = "CD4.Q_th1",
    group2 = "CD4.Q_th17",
    signatures = c(
      "Th1",
      "Th17"
    ),
    filename = "QTh1_vs_QTh17"
  ),

  "CD4.Q_th2_vs_CD4.Q_th17" = list(
    group1 = "CD4.Q_th2",
    group2 = "CD4.Q_th17",
    signatures = c(
      "Th2",
      "Th17"
    ),
    filename = "QTh2_vs_QTh17"
  )
)


for (comp in names(comparison_info)) {

  info <- comparison_info[[comp]]

  de <- q_dge %>%
    filter(
      comparison == comp
    )

  label_df <- de %>%
    filter(
      signature %in% info$signatures
    )

  p <- ggplot(
    de,
    aes(
      x = avg_log2FC,
      y = neglog10_padj
    )
  ) +

    geom_point(
      data = de %>%
        filter(
          !signature %in% info$signatures
        ),
      color = "lightgray",
      size = 0.7,
      alpha = 0.5
    ) +

    geom_point(
      data = de %>%
        filter(
          signature %in% info$signatures
        ),
      aes(
        color = signature
      ),
      size = 1.4,
      alpha = 0.9
    ) +

    geom_text_repel(
      data = label_df,
      aes(
        label = gene,
        color = signature
      ),
      size = 2.5,
      box.padding = 0.35,
      point.padding = 0.2,
      min.segment.length = 0,
      max.overlaps = Inf,
      seed = 1,
      show.legend = FALSE
    ) +

    scale_color_manual(
      values = signature_colors
    ) +

    geom_vline(
      xintercept = 0,
      linetype = "dashed",
      linewidth = 0.25
    ) +

    labs(
      x = paste0(
        "log2FC (",
        info$group1,
        " vs ",
        info$group2,
        ")"
      ),
      y = "-log10 adjusted p-value",
      color = NULL,
      title = paste0(
        info$group1,
        " vs ",
        info$group2
      )
    ) +

    theme_classic(
      base_size = 9
    ) +

    theme(
      legend.position = "right"
    )


  print(p)

}
```
# Extended Data Figure 5

## Polyfunctional cell frequencies reflect random assortment of Th programs at low expression levels

### Fig S5
```{r S5 - kappa heatmaps}
# run_kappa_heatmap.R
#
# Standalone script: plots the kappa joint-tail heatmap for all pairwise
# TH program comparisons.
#
#   kappa(x, y) = P(X > x, Y > y)          <- numerator:   empirical joint tail
#                 ──────────────────
#                 P(X > x) * P(Y > y)       <- denominator: independence baseline
#
#   kappa < 1  =>  joint tail DEPLETION   (programs mutually exclusive at extremes)
#   kappa = 1  =>  independence
#   kappa > 1  =>  joint tail ENRICHMENT  (programs co-occur at extremes)
#
# Output: one PDF per pair in figure/kappa_heatmap/

source("code/tail_dependence_functions.R")

DATA_PATH <- "data/metadata_with_modulescore_annotation.csv"

N_GRID  <- 80
V_FIXED <- 0.80

PROGRAMS <- c(
  "TH1_Score1",
  "TH2_Score1",
  "TH17_Score1",
  "TFH_TCHRON_Score1"
)

metadata <- read.csv(DATA_PATH)

short_label <- function(nm) {
  sub("_Score1$", "", nm)
}

pretty_module_label <- function(score_name) {
  lab <- short_label(score_name)

  if (toupper(lab) == "TH17") {
    return("Th17-like module score")
  }

  if (toupper(lab) == "TH1") {
    return("Th1-like module score")
  }

  if (toupper(lab) == "TH2") {
    return("Th2-like module score")
  }

  if (toupper(lab) == "TFH_TCHRON") {
    return("Tfh/Tchron-like module score")
  }

  sprintf("%s-like module score", lab)
}

plot_kappa_heatmap <- function(
  x,
  y,
  x_label,
  y_label,
  n_grid = 80,
  v_fixed = 0.80,
  zlim_kappa = NULL
) {
  plot_kappa_heatmap_square(
    x = x,
    y = y,
    x_label = x_label,
    y_label = y_label,
    n_grid = n_grid,
    v_fixed = v_fixed,
    zlim_kappa = zlim_kappa,
    main = ""
  )
}

pairs <- combn(
  PROGRAMS,
  2,
  simplify = FALSE
)

for (pair in pairs) {

  x_name <- pair[1]
  y_name <- pair[2]

  x_axis <- pretty_module_label(x_name)
  y_axis <- pretty_module_label(y_name)

  plot_kappa_heatmap(
    x = metadata[[x_name]],
    y = metadata[[y_name]],
    x_label = x_axis,
    y_label = y_axis,
    n_grid = N_GRID,
    v_fixed = V_FIXED
  )
}

```
# Extended Data Figure 6

## Memory CD4+ T cells

### Fig S6a-b

```{r S6a/b Feature plots of Tbx21 and Ifng in SLOs during acute listeria infection and memory, fig.width=8, fig.height. = 8}
MyFeaturePlotHighlight <- function (seurat_object, feature, cells_to_highlight, umap_to_plot = "umap", 
    assay = "RNA", colors = viridis::viridis(10), title = "", 
    pixels = c(512, 512), highlight_size = 1, highlight_alpha = 1, 
    xlim = NULL, ylim = NULL, background_alpha = 0.1) 
{
    dim_data <- seurat_object[[umap_to_plot]]@cell.embeddings
    dim1 <- dim_data[, 1]
    dim2 <- dim_data[, 2]
    tmp_all <- data.frame(seurat_object@meta.data, dim1 = dim1, 
        dim2 = dim2)
    p_base <- ggplot2::ggplot(tmp_all) + scattermore::geom_scattermore(ggplot2::aes(x = dim1, 
        y = dim2), color = "grey", pointsize = 1, alpha = background_alpha, 
        pixels = pixels)
    feature_expr <- Seurat::GetAssayData(seurat_object, assay = assay, 
        layer = "data")[feature, ]
    tmp_highlight <- tmp_all[cells_to_highlight, ]
    tmp_highlight$feature_value <- feature_expr[cells_to_highlight]
    p_final <- p_base + ggplot2::geom_point(data = tmp_highlight, 
        ggplot2::aes(x = dim1, y = dim2, color = feature_value), 
        size = highlight_size, alpha = highlight_alpha) + ggplot2::scale_color_gradientn(colours = colors, 
        name = feature) + ggplot2::theme_void() + NoGrid() + 
        ggplot2::labs(title = title, x = paste0(umap_to_plot, 
            "_1"), y = paste0(umap_to_plot, "_2"))
    if (!is.null(xlim) || !is.null(ylim)) {
        p_final <- p_final + ggplot2::coord_cartesian(xlim = xlim, 
            ylim = ylim)
    }
    return(p_final)
}

feature <- "Tbx21"
listeria_d610 <- subset(listeria_object, subset = condition_detailed %in% c("Lmonocytogenes_D6", "Lmonocytogenes_D10"))
cells_to_highlight <- colnames(listeria_d610)
MyFeaturePlotHighlight(
  seurat_object      = cd4_object,
  feature            = feature,
  cells_to_highlight = cells_to_highlight,
  umap_to_plot       = "mde_incremental",
  assay              = "RNA",
  colors = c("lightgrey", "blue"),
  highlight_size = 3.0,
  background_alpha   = 0.25, 
  title = "Tbx21 in Listeria Infection d6/10"
)

feature <- "Ifng"
MyFeaturePlotHighlight(
  seurat_object      = cd4_object,
  feature            = feature,
  cells_to_highlight = cells_to_highlight,
  umap_to_plot       = "mde_incremental",
  assay              = "RNA",
  colors = c("lightgrey", "blue"),
  highlight_size = 3.0,
  background_alpha   = 0.25, 
  title = "Ifng in Listeria Infection d6/10"
)

#d253060
feature <- "Tbx21"
listeria_d253060 <- subset(listeria_object, subset = condition_detailed %in% c("Lmonocytogenes_D25", "Lmonocytogenes_D30", "Lmonocytogenes_D60"))
cells_to_highlight <- colnames(listeria_d253060)
MyFeaturePlotHighlight(
  seurat_object      = cd4_object,
  feature            = feature,
  cells_to_highlight = cells_to_highlight,
  umap_to_plot       = "mde_incremental",
  assay              = "RNA",
  colors = c("lightgrey", "blue"),
  highlight_size = 3.0,
  background_alpha   = 0.25, 
  title = "Tbx21 in Listeria Infection d25/30/60"
)

feature <- "Ifng"
MyFeaturePlotHighlight(
  seurat_object      = cd4_object,
  feature            = feature,
  cells_to_highlight = cells_to_highlight,
  umap_to_plot       = "mde_incremental",
  assay              = "RNA",
  colors = c("lightgrey", "blue"),
  highlight_size = 3.0,
  background_alpha   = 0.25, 
  title = "Ifng in Listeria Infection d25/30/60"
)
```
### Fig S6c
```{r S6c - david}

```
### Fig S6e
```{r s6e mdes of smarta cells in SLOs at day 7 and d21, fig.width=8, fig.height=8}
clusters_keep <- c(
  "CD4.Q", "CD4.G", "CD4.H", "CD4.I",
  "CD4.R", "CD4.S", "CD4.T", "CD4.U",
  "CD4.X", "CD4.Y"
)
modulescores <- FetchData(
  cd4_object,
  vars = c(
    "TH1_Score1",
    "TH2_Score1",
    "TH17_Score1",
    "TFH_TCHRON_Score1",
    "annotation_level2",
    "organ_simplified",
    "condition_detailed"
  )
) %>%
  rownames_to_column("cell") %>%
  mutate(is_smarta = cell %in% smarta_ids) %>%
  filter(annotation_level2 %in% clusters_keep)


smarta_meta <- modulescores %>%
  filter(is_smarta)

smarta_d7_slo <- modulescores %>%
  filter(
    is_smarta,
    organ_simplified %in% c("spleen", "SLO"),
    condition_detailed %in% c("LCMVarm_D7")
  )

#th1 vs tfh
quadrant_pct <- smarta_d7_slo %>%
  mutate(
    quadrant = case_when(
      TFH_TCHRON_Score1 >= 0 & TH1_Score1 >= 0 ~ "Q1 (TFH_TCHRON+TH1+)",
      TFH_TCHRON_Score1 <  0 & TH1_Score1 >= 0 ~ "Q2 (TFH_TCHRON-TH1+)",
      TFH_TCHRON_Score1 <  0 & TH1_Score1 <  0 ~ "Q3 (TFH_TCHRON-TH1-)",
      TFH_TCHRON_Score1 >= 0 & TH1_Score1 <  0 ~ "Q4 (TFH_TCHRON+TH1-)"
    )
  ) %>%
  count(quadrant) %>%
  mutate(
    pct = n / sum(n) * 100,
    label = sprintf("%.1f%%", pct)
  )

quadrant_coords <- data.frame(
  quadrant = c("Q1 (TFH_TCHRON+TH1+)", "Q2 (TFH_TCHRON-TH1+)", "Q3 (TFH_TCHRON-TH1-)", "Q4 (TFH_TCHRON+TH1-)"),
  x = c(1.2, -0.1, -0.1, 1.2),
  y = c(1.0, 1.3, -0.2, -0.2)
)

quadrant_pct <- left_join(quadrant_pct, quadrant_coords, by = "quadrant")

mean_smarta <- smarta_d7_slo %>%
  group_by(annotation_level2) %>%
  summarise(
    TFH_TCHRON = mean(TFH_TCHRON_Score1, na.rm = TRUE),
    TH1 = mean(TH1_Score1, na.rm = TRUE),
    .groups = "drop"
  )

ggplot() +
  geom_point_rast(
    data = modulescores,
    aes(x = TFH_TCHRON_Score1, y = TH1_Score1),
    color = "grey80",
    size = 0.35,
    alpha = 0.25,
    raster.dpi = 200
  ) +
  geom_point_rast(
    data = smarta_d7_slo,
    aes(x = TFH_TCHRON_Score1, y = TH1_Score1, color = annotation_level2),
    size = 3,
    alpha = 0.9,
    raster.dpi = 200
  ) +
  geom_point(
    data = mean_smarta,
    aes(x = TFH_TCHRON, y = TH1, fill = annotation_level2),
    shape = 21,
    color = "black",
    size = 4,
    stroke = 1
  ) +
  geom_text(
    data = mean_smarta,
    aes(x = TFH_TCHRON, y = TH1, label = annotation_level2),
    color = "black",
    vjust = -1,
    size = 4
  ) +
  geom_hline(
    yintercept = 0,
    color = "red",
    linetype = "dashed",
    linewidth = 0.4
  ) +
  geom_vline(
    xintercept = 0,
    color = "red",
    linetype = "dashed",
    linewidth = 0.4
  ) +
  geom_text(
    data = quadrant_pct,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    color = "black",
    fontface = "bold",
    size = 4
  ) +
  theme_classic() +
  labs(
    x = "Tfh_Tchron-like module score",
    y = "Th1-like module score",
    color = "Cluster",
    fill = "Cluster"
  ) +
  coord_cartesian(xlim = c(-0.4, 1.5), ylim = c(-0.5, 1.5)) +
  scale_color_manual(values = palette_vector) +
  scale_fill_manual(values = palette_vector)


#th2 vs th17
quadrant_pct <- smarta_d7_slo %>%
  mutate(
    quadrant = case_when(
      TH17_Score1 >= 0.25 & TH2_Score1 >= 0.25 ~ "Q1 (TH17+TH2+)",
      TH17_Score1 <  0.25 & TH2_Score1 >= 0.25 ~ "Q2 (TH17-TH2+)",
      TH17_Score1 <  0.25 & TH2_Score1 <  0.25 ~ "Q3 (TH17-TH2-)",
      TH17_Score1 >= 0.25 & TH2_Score1 <  0.25 ~ "Q4 (TH17+TH2-)"
    )
  ) %>%
  count(quadrant) %>%
  mutate(
    pct = n / sum(n) * 100,
    label = sprintf("%.1f%%", pct)
  )

quadrant_coords <- data.frame(
  quadrant = c("Q1 (TH17+TH2+)", "Q2 (TH17-TH2+)", "Q3 (TH17-TH2-)", "Q4 (TH17+TH2-)"),
  x = c(1.2, -0.1, -0.1, 1.2),
  y = c(1.0, 1.3, -0.2, -0.2)
)

quadrant_pct <- left_join(quadrant_pct, quadrant_coords, by = "quadrant")

mean_smarta <- smarta_d7_slo %>%
  group_by(annotation_level2) %>%
  summarise(
    TH17 = mean(TH17_Score1, na.rm = TRUE),
    TH2 = mean(TH2_Score1, na.rm = TRUE),
    .groups = "drop"
  )

ggplot() +
  geom_point_rast(
    data = modulescores,
    aes(x = TH17_Score1, y = TH2_Score1),
    color = "grey80",
    size = 0.35,
    alpha = 0.25,
    raster.dpi = 200
  ) +
  geom_point_rast(
    data = smarta_d7_slo,
    aes(x = TH17_Score1, y = TH2_Score1, color = annotation_level2),
    size = 3,
    alpha = 0.9,
    raster.dpi = 200
  ) +
  geom_point(
    data = mean_smarta,
    aes(x = TH17, y = TH2, fill = annotation_level2),
    shape = 21,
    color = "black",
    size = 4,
    stroke = 1
  ) +
  geom_text(
    data = mean_smarta,
    aes(x = TH17, y = TH2, label = annotation_level2),
    color = "black",
    vjust = -1,
    size = 4
  ) +
  geom_hline(
    yintercept = 0.25,
    color = "red",
    linetype = "dashed",
    linewidth = 0.4
  ) +
  geom_vline(
    xintercept = 0.25,
    color = "red",
    linetype = "dashed",
    linewidth = 0.4
  ) +
  geom_text(
    data = quadrant_pct,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    color = "black",
    fontface = "bold",
    size = 4
  ) +
  theme_classic() +
  labs(
    x = "Th17-like module score",
    y = "Th2-like module score",
    color = "Cluster",
    fill = "Cluster"
  ) +
  coord_cartesian(xlim = c(-0.4, 1.5), ylim = c(-0.5, 1.5)) +
  scale_color_manual(values = palette_vector) +
  scale_fill_manual(values = palette_vector)

#memory
smarta_d3060_slo <- modulescores %>%
  filter(
    is_smarta,
    organ_simplified %in% c("spleen", "LN"),
    condition_detailed %in% c("LCMVarm_D30", "LCMVarm_D60")
  )



quadrant_pct <- smarta_d3060_slo %>%
  mutate(
    quadrant = case_when(
      TH17_Score1 >= 0.25 & TH2_Score1 >= 0.25 ~ "Q1 (TH17+TH2+)",
      TH17_Score1 <  0.25 & TH2_Score1 >= 0.25 ~ "Q2 (TH17-TH2+)",
      TH17_Score1 <  0.25 & TH2_Score1 <  0.25 ~ "Q3 (TH17-TH2-)",
      TH17_Score1 >= 0.25 & TH2_Score1 <  0.25 ~ "Q4 (TH17+TH2-)"
    )
  ) %>%
  count(quadrant) %>%
  mutate(
    pct = n / sum(n) * 100,
    label = sprintf("%.1f%%", pct)
  )

quadrant_coords <- data.frame(
  quadrant = c("Q1 (TH17+TH2+)", "Q2 (TH17-TH2+)", "Q3 (TH17-TH2-)", "Q4 (TH17+TH2-)"),
  x = c(1.2, -0.1, -0.1, 1.2),
  y = c(1.0, 1.3, -0.2, -0.2)
)

quadrant_pct <- left_join(quadrant_pct, quadrant_coords, by = "quadrant")

mean_smarta <- smarta_d3060_slo %>%
  group_by(annotation_level2) %>%
  summarise(
    TH17 = mean(TH17_Score1, na.rm = TRUE),
    TH2 = mean(TH2_Score1, na.rm = TRUE),
    .groups = "drop"
  )


ggplot() +
  geom_point_rast(
    data = modulescores,
    aes(x = TH17_Score1, y = TH2_Score1),
    color = "grey80",
    size = 0.35,
    alpha = 0.25,
    raster.dpi = 200
  ) +
  geom_point_rast(
    data = smarta_d3060_slo,
    aes(x = TH17_Score1, y = TH2_Score1, color = annotation_level2),
    size = 3,
    alpha = 0.9,
    raster.dpi = 200
  ) +
  geom_point(
    data = mean_smarta,
    aes(x = TH17, y = TH2, fill = annotation_level2),
    shape = 21,
    color = "black",
    size = 4,
    stroke = 1
  ) +
  geom_text(
    data = mean_smarta,
    aes(x = TH17, y = TH2, label = annotation_level2),
    color = "black",
    vjust = -1,
    size = 4
  ) +
  geom_hline(
    yintercept = 0.25,
    color = "red",
    linetype = "dashed",
    linewidth = 0.4
  ) +
  geom_vline(
    xintercept = 0.25,
    color = "red",
    linetype = "dashed",
    linewidth = 0.4
  ) +
  geom_text(
    data = quadrant_pct,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    color = "black",
    fontface = "bold",
    size = 4
  ) +
  theme_classic() +
  labs(
    x = "Th17-like module score",
    y = "Th2-like module score",
    color = "Cluster",
    fill = "Cluster"
  ) +
  coord_cartesian(xlim = c(-0.4, 1.5), ylim = c(-0.5, 1.5)) +
  scale_color_manual(values = palette_vector) +
  scale_fill_manual(values = palette_vector)



#th1 vs tfhtchron
quadrant_pct <- smarta_d3060_slo %>%
  mutate(
    quadrant = case_when(
      TFH_TCHRON_Score1 >= 0 & TH1_Score1 >= 0 ~ "Q1 (TFH_TCHRON+TH1+)",
      TFH_TCHRON_Score1 <  0 & TH1_Score1 >= 0 ~ "Q2 (TFH_TCHRON-TH1+)",
      TFH_TCHRON_Score1 <  0 & TH1_Score1 <  0 ~ "Q3 (TFH_TCHRON-TH1-)",
      TFH_TCHRON_Score1 >= 0 & TH1_Score1 <  0 ~ "Q4 (TFH_TCHRON+TH1-)"
    )
  ) %>%
  count(quadrant) %>%
  mutate(
    pct = n / sum(n) * 100,
    label = sprintf("%.1f%%", pct)
  )

quadrant_coords <- data.frame(
  quadrant = c("Q1 (TFH_TCHRON+TH1+)", "Q2 (TFH_TCHRON-TH1+)", "Q3 (TFH_TCHRON-TH1-)", "Q4 (TFH_TCHRON+TH1-)"),
  x = c(1.2, -0.1, -0.1, 1.2),
  y = c(1.0, 1.3, -0.2, -0.2)
)

quadrant_pct <- left_join(quadrant_pct, quadrant_coords, by = "quadrant")

mean_smarta <- smarta_d3060_slo %>%
  group_by(annotation_level2) %>%
  summarise(
    TFH_TCHRON = mean(TFH_TCHRON_Score1, na.rm = TRUE),
    TH1 = mean(TH1_Score1, na.rm = TRUE),
    .groups = "drop"
  )


ggplot() +
  geom_point_rast(
    data = modulescores,
    aes(x = TFH_TCHRON_Score1, y = TH1_Score1),
    color = "grey80",
    size = 0.35,
    alpha = 0.25,
    raster.dpi = 200
  ) +
  geom_point_rast(
    data = smarta_d3060_slo,
    aes(x = TFH_TCHRON_Score1, y = TH1_Score1, color = annotation_level2),
    size = 3,
    alpha = 0.9,
    raster.dpi = 200
  ) +
  geom_point(
    data = mean_smarta,
    aes(x = TFH_TCHRON, y = TH1, fill = annotation_level2),
    shape = 21,
    color = "black",
    size = 4,
    stroke = 1
  ) +
  geom_text(
    data = mean_smarta,
    aes(x = TFH_TCHRON, y = TH1, label = annotation_level2),
    color = "black",
    vjust = -1,
    size = 4
  ) +
  geom_hline(
    yintercept = 0,
    color = "red",
    linetype = "dashed",
    linewidth = 0.4
  ) +
  geom_vline(
    xintercept = 0,
    color = "red",
    linetype = "dashed",
    linewidth = 0.4
  ) +
  geom_text(
    data = quadrant_pct,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    color = "black",
    fontface = "bold",
    size = 4
  ) +
  theme_classic() +
  labs(
    x = "Tfh_Tchron-like module score",
    y = "Th1-like module score",
    color = "Cluster",
    fill = "Cluster"
  ) +
  coord_cartesian(xlim = c(-0.4, 1.5), ylim = c(-0.5, 1.5)) +
  scale_color_manual(values = palette_vector) +
  scale_fill_manual(values = palette_vector)


```
# Extended Data Figure 7

## Resting CD4+ T cell clusters

### Fig S7a
```{r S7a }

```

### Fig S7b
```{r S7b - external dataset mapping for Sajani et al, fig.width = 8, fig.height=8}
obj <- NormalizeData(obj)
obj <- FindVariableFeatures(obj)
obj <- ScaleData(obj)
obj <- RunPCA(obj)

obj$immgen_clusters <- obj$level2_final

obj <- FindNeighbors(obj, dims = 1:30)
obj <- RunUMAP(obj, dims = 1:30)

for (res in c(0.2, 0.3, 0.4, 0.5, 0.6)) {
  obj <- FindClusters(obj, resolution = res)

  cat("\nresolution =", res, "\n")
  print(table(Idents(obj)))
}

obj <- FindClusters(obj, resolution = 0.2)
obj$clusters_0.2 <- Idents(obj)

obj <- RenameIdents(
  obj,
  "0" = "TUND",
  "1" = "TIFN",
  "2" = "TQ",
  "3" = "TMEM",
  "4" = "TTCR"
)

obj$cluster_named <- Idents(obj)

DimPlot(
  obj,
  group.by = "cluster_named",
  label = TRUE
)

ref_embedding <- as.data.frame(
  Embeddings(
    cd4_object,
    reduction = "mde_incremental"
  )
)

colnames(ref_embedding) <- c(
  "UMAP_1",
  "UMAP_2"
)

query_embedding <- as.data.frame(
  Embeddings(
    obj,
    reduction = "mde_incremental_level2"
  )
)

colnames(query_embedding) <- c(
  "UMAP_1",
  "UMAP_2"
)

query_embedding$cluster_named <- as.character(
  obj$cluster_named
)

cluster_colors <- c(
  "TUND" = "#E69F00",
  "TIFN" = "#56B4E9",
  "TQ" = "#009E73",
  "TMEM" = "#CC79A7",
  "TTCR" = "#0072B2"
)

p <- ggplot() +
  geom_point(
    data = ref_embedding,
    aes(
      x = UMAP_1,
      y = UMAP_2
    ),
    color = "lightgray",
    size = 0.3,
    alpha = 0.5
  ) +
  geom_point(
    data = query_embedding,
    aes(
      x = UMAP_1,
      y = UMAP_2,
      color = cluster_named
    ),
    size = 0.3,
    alpha = 0.8
  ) +
  theme_classic() +
  labs(
    title = "Query Clusters on Reference Embedding",
    color = "Cluster"
  ) +
  guides(
    color = guide_legend(
      override.aes = list(size = 3)
    )
  ) +
  scale_color_manual(
    values = cluster_colors
  )

print(p)
```
### Fig S7c
```{r S7c - heatmap comparing sajani et al to igt cd4 resting clusters}
Idents(obj) <- obj$clusters_0.2

markers <- FindAllMarkers(
  obj,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

top10 <- markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 10) %>%
  arrange(cluster, desc(avg_log2FC))

ordered_genes <- top10$gene

avg_exp <- AverageExpression(
  obj,
  features = ordered_genes,
  group.by = "clusters_0.2",
  assays = "RNA"
)$RNA

heat_mat <- t(scale(t(avg_exp)))

heat_mat_ordered <- heat_mat[ordered_genes, ]

annotation_row <- data.frame(
  cluster = top10$cluster
)
rownames(annotation_row) <- ordered_genes

cluster_sizes <- table(top10$cluster)
gaps_row <- cumsum(cluster_sizes)

pheatmap(
  heat_mat_ordered,
  color = colorRampPalette(c("darkblue", "white", "red"))(100),
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_row = annotation_row,
  fontsize_col = 10,
  cellwidth = 12,
  cellheight = 12,
  border_color = "lightgray"
)

Idents(obj) <- obj$clusters_0.2

obj <- RenameIdents(
  obj,
  "0" = "TUND",
  "1" = "TIFN",
  "2" = "TQ",
  "3" = "TMEM",
  "4" = "TTCR"
)


obj$cluster_named <- Idents(obj)


DimPlot(
  obj,
  reduction = "mde_incremental_level2",
  group.by = "cluster_named",
  label = TRUE
)


df_out <- data.frame(
  Cell_ID = colnames(obj),
  batch = "b6spleenb6ln",
  new_annotation = obj$cluster_named
)

write.csv(df_out, "cell_annotations.csv", row.names = FALSE)
```
# Extended Data Figure 8

## immgenT CD4 flow cytometry panel examples

### Fig S8a
```{r S8a hgihlighting cd44+cd62l+ cells, fig.width = 8, fig.height=8}
cd62l_feat <- "CD62L"
cd44_feat <- "CD44"

cd62l_thresh <- 5
cd44_thresh <- 5


cd4_cite$is_cd62l_cd44 <-
  (adt[cd62l_feat, ]  > cd62l_thresh) &
  (adt[cd44_feat, ] > cd44_thresh)

p <- MyDimPlotHighlightDensity(
  seurat_object = cd4_cite,
  umap_to_plot = "mde_incremental",
  group.by = "is_cd62l_cd44",
  split.by = NULL,
  raster = FALSE,
  highlight_size = 0.25,
  highlight_alpha = 1.0,
  cols = viridis(100)
)
print(p)
```
### Fig S8b
```{r s8b feature scatter plot of klrg1 x cd62l + corresponding mde, fig.width=8, fig.height=8}
cd4_cite$not_cl6 <- cd4_cite$annotation_level2 != "CD4_cl6"
p <- MyFeatureScatter(
  so = cd4_cite,
  group.by = "not_cl6",
  feature1 = "CD62L",
  feature2 = "KLRG1",
  raster = TRUE
)
print(p)

cd62l_feat <- "CD62L"
klrg1_feat <- "KLRG1"

klrg1_thresh <- 5
cd62l_thresh <- 5


cd4_cite$is_cd62l_klrg1 <-
  (adt[cd62l_feat, ]  > cd62l_thresh) &
  (adt[klrg1_feat, ] > klrg1_thresh)

p <- MyDimPlotHighlightDensity(
  seurat_object = cd4_cite,
  umap_to_plot = "mde_incremental",
  group.by = "is_cd62l_klrg1",
  split.by = NULL,
  raster = FALSE,
  highlight_size = 0.25,
  highlight_alpha = 1.0,
  cols = viridis(100)
)
print(p)
```


